---
title: "Blackbird Model Performance Exploration"
author: "Michele Wiseman"
date: "2024-01-30"
output:
  pdf_document: default
  html_document: default
toc: true
toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Help here: <https://emdelponte.github.io/paper-estimate-app/code.html>

Ideas: - Look at rater severity rate compared done on two separate days (human precision estimates) - Look at saliency ratings vs human ratings

# Set up

```{r load packges, warning=FALSE, echo=FALSE}
library(ggplot2)
#install.packages(c("tidyverse"))

library(ggtext)
library(GGally)
library(lme4)
library(tidyverse)
library(janitor)
library(plotly)
library(readxl)
library(reshape2)
library(patchwork)
library(psych)
library(tidyr)
#library(epiR)
library(agricolae)
library(ggridges)
library(dplyr)
library(lubridate)
library(readr)
library(purrr)
library(vroom)
library(data.table)
library(here)

```

## Set working directory

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/Intel User/Desktop/blackbird_scripts/r")
```


## Import data
```{r import data}

getwd() # this should be wherever your Rmd is... the package "here" handles that

correlation_matrix_human_v_model <- read_excel("../spreadsheets/correlation_matrix_human_v_model_extended_all.xlsx")

# drop time stamp column
correlation_matrix_human_v_model <- correlation_matrix_human_v_model %>%
  select(-timestamp)

#View(correlation_matrix_human_v_model)
```

## Concatenate blackbird data
```{r concat blackbird results}
find_csv_files <- function(root_path) {
  list.files(root_path, pattern = "\\.csv$", recursive = TRUE, full.names = TRUE)
}


remove_column_from_csv <- function(file_path, column_name) {
  df <- read_csv(file_path, show_col_types = FALSE)

  # convert timestamp BEFORE concat
  if ("timestamp" %in% names(df)) {
    ts_chr <- as.character(df$timestamp)

    # matches "1/24/2026 17:32"
    parsed <- mdy_hm(ts_chr, tz = "UTC", quiet = TRUE)

    # fallback if some rows include seconds like "1/24/2026 17:32:10"
    if (any(is.na(parsed)) && any(grepl(":\\d\\d$", ts_chr))) {
      parsed2 <- mdy_hms(ts_chr, tz = "UTC", quiet = TRUE)
      parsed[is.na(parsed)] <- parsed2[is.na(parsed)]
    }

    df$timestamp <- parsed
  }

  # remove unwanted column (works even for name "0")
  if (column_name %in% names(df)) {
    df <- df[, names(df) != column_name, drop = FALSE]
  }

  df %>% filter(!if_all(everything(), is.na))
}

main <- function() {
  root_paths <- c("C:/Users/Intel User/Desktop/blackbird_scripts/data/Results")
  column_to_remove <- "0"
  output_file <- "../spreadsheets/grid_search_concatenated_results.csv"

  all_csv_files <- unlist(lapply(root_paths, find_csv_files))
  message("Found ", length(all_csv_files), " CSV files")

  modified_dfs <- map(all_csv_files, ~ remove_column_from_csv(.x, column_to_remove))
  concatenated_df <- bind_rows(modified_dfs)

  write_csv(concatenated_df, output_file)
  message("Concatenated file created: ", output_file)

  concatenated_df
}

concatenated_df <- main()
head(concatenated_df)

# if up_th exists, put those values in infected threshold and remove that column
concatenated_df <- concatenated_df %>%
  mutate(
    infected_threshold = ifelse(!is.na(up_th), up_th, infected_threshold)
  ) %>%
  select(-up_th)

concatenated_df <- concatenated_df %>%
  mutate(
    healthy_threshold = ifelse(!is.na(down_th), down_th, healthy_threshold)
  ) %>%
  select(-down_th)


```
## Data tidying
```{r data tidying}

# rename infected_threshold colname to up_th
colnames(concatenated_df)[which(names(concatenated_df) == "infected_threshold")] <- "up_th"

# rename healthy_threshold colname to down_th
colnames(concatenated_df)[which(names(concatenated_df) == "healthy_threshold")] <- "down_th"

# change model_timestamp to identifier
colnames(concatenated_df)[which(names(concatenated_df) == "model_timestamp")] <- "identifier"

#remove timestamp column
concatenated_df <- concatenated_df %>%
  select(-timestamp)

# change classes from 2 to 3 if model identifier is Jan22_10-21-00_2026
concatenated_df <- concatenated_df %>%
  mutate(classes = ifelse(identifier == "Jan22_10-21-00_2026", 3, classes)) %>%
  mutate(classes = ifelse(identifier == "Jan26_23-15-35_2026", 3, classes)) %>%
  mutate(classes = ifelse(identifier == "Jan29_22-02-46_2026", 3, classes))


concatenated_df <- concatenated_df %>%
  select(-time_elapsed)


# deduplicate any duplicate rows
concatenated_df <- concatenated_df %>%  distinct()


# add column model_type between timestamp and identifier. populate with "computer_vision"
concatenated_df <- concatenated_df %>%
  mutate(model_type = "computer_vision", .before = identifier)

cols_to_keep <- names(correlation_matrix_human_v_model)

concatenated_df_trimmed <- concatenated_df %>%
  select(any_of(c(cols_to_keep, "inf_gate", "spor_th","GradCAM_sr2", "DeepLift_sr2")))

combined <- bind_rows(
  correlation_matrix_human_v_model %>% mutate(across(everything(), as.character)),
  concatenated_df_trimmed %>% mutate(across(everything(), as.character))
)

# populate percent_healthy, percent_hyphae, percent_conidiophores, and percent_infected

combined <- combined %>%
  mutate(
    # turn "-" and "" into NA for the patch columns, then parse as numbers
    across(
      c(clear_patches, hyphal_patches, conidiophore_patches),
      ~ readr::parse_number(na_if(na_if(as.character(.x), "-"), ""))
    ),
    total_patches = clear_patches + hyphal_patches + conidiophore_patches,

    percent_infected = if_else(
      model_type == "computer_vision" & !is.na(total_patches) & total_patches > 0,
      100 * ((hyphal_patches + conidiophore_patches) / total_patches),
      readr::parse_number(na_if(na_if(as.character(percent_infected), "-"), ""))
    ),

    percent_healthy = if_else(
      model_type == "computer_vision" & !is.na(total_patches) & total_patches > 0,
      100 - percent_infected,
      readr::parse_number(na_if(na_if(as.character(percent_healthy), "-"), ""))
    ),

    percent_hyphae = if_else(
      model_type == "computer_vision" & !is.na(total_patches) & total_patches > 0,
      100 * (hyphal_patches / total_patches),
      readr::parse_number(na_if(na_if(as.character(percent_hyphae), "-"), ""))
    ),

    percent_conidiophores = if_else(
      model_type == "computer_vision" & !is.na(total_patches) & total_patches > 0,
      100 * (conidiophore_patches / total_patches),
      readr::parse_number(na_if(na_if(as.character(percent_conidiophores), "-"), ""))
    )
  ) %>%
  select(-total_patches)

```

## Skip the chunk below unless running grid search
I've set it to eval=FALSE so it won't run -- change it back if you need it. 

```{r pivoting table, eval=FALSE}

# Count unique values for identifier
n_unique_filenames <- combined %>%
  dplyr::group_by(identifier) %>%
  dplyr::summarize(unique_filenames = n_distinct(filename))

View(n_unique_filenames)

# select only file names that are found in Feb14_15-53-04_2024 model across all models
unique_filenames <- combined %>%
 # dplyr::filter(identifier == "Jan22_10-21-00_2026") %>%
  dplyr::select(filename) %>%
  dplyr::distinct()

# filter combined to only include these filenames
common_files <- combined %>%
  dplyr::filter(filename %in% unique_filenames$filename)


# are there any identifiers who dont have all of these filenames?
n_unique_filenames <- common_files %>%
  dplyr::group_by(identifier) %>%
  dplyr::summarize(unique_filenames = n_distinct(filename))

# I want to remove any rows that don't have a complete set of ratings from all raters
all_ids <- sort(unique(common_files$identifier))

complete_files <- common_files %>%
  dplyr::distinct(filename, identifier) %>%   # avoid double-counting
  dplyr::count(filename, name = "n_ids") %>%
  dplyr::filter(n_ids == length(all_ids)) %>%
  dplyr::pull(filename)

common_files_complete <- common_files %>%
  dplyr::filter(filename %in% complete_files)

```

## More data tidying
```{r data tidying rnd 2}

# select only the relevant columns
correlation_matrix_human_v_model_filtered_select <- combined %>%
        dplyr::select(filename, identifier, model_type, tray, classes, up_th, down_th, spor_th, inf_gate, sal_threshold, severity_rate_patch) %>%
        dplyr::filter(!is.na(severity_rate_patch))


# if the filename doesnt end in "5dpi" or "10dpi", then append this information to the filename from the tray column
correlation_matrix_human_v_model_filtered_select <- correlation_matrix_human_v_model_filtered_select %>%
  dplyr::mutate(filename = dplyr::case_when(
    grepl("5dpi", filename) ~ filename,
    grepl("10dpi", filename) ~ filename,
    TRUE ~ paste0(filename, "_", tray)
  ))

correlation_matrix_human_v_model_filtered_select2 <- correlation_matrix_human_v_model_filtered_select %>%
  dplyr::mutate(
    up_th_num = suppressWarnings(as.numeric(up_th)),
    identifier = dplyr::if_else(
      !is.na(up_th_num),                      # only if it's numeric
      paste0(identifier, "_up", up_th),         # append original value
      identifier                              # otherwise leave unchanged
    )
  ) %>%
  dplyr::mutate(
    down_th_num = suppressWarnings(as.numeric(down_th)),
    identifier = dplyr::if_else(
      !is.na(down_th_num),                      # only if it's numeric
      paste0(identifier, "_dwn", down_th),         # append original value
      identifier                              # otherwise leave unchanged
      )) %>%
    dplyr::mutate(
    inf_gate_num = suppressWarnings(as.numeric(inf_gate)),
    identifier = dplyr::if_else(
      !is.na(inf_gate_num),                      # only if it's numeric
      paste0(identifier, "_infg", inf_gate),         # append original value
      identifier                              # otherwise leave unchanged
      )) %>%
    dplyr::mutate(
      spor_th_num = suppressWarnings(as.numeric(spor_th)),
      identifier = dplyr::if_else(
      !is.na(spor_th_num),                      # only if it's numeric
      paste0(identifier, "_spth", spor_th),         # append original value
      identifier                              # otherwise leave unchanged
      )) %>%
    dplyr::mutate(
      sal_th_num = suppressWarnings(as.numeric(sal_threshold)),
      identifier = dplyr::if_else(
        !is.na(sal_th_num),                      # only if it's numeric
        paste0(identifier, "_salth", sal_threshold),         # append original value
        identifier                             # otherwise leave unchanged
        )) %>%
dplyr::select(c(-up_th_num, -down_th_num, -inf_gate_num, -spor_th_num, -sal_th_num))

unique(correlation_matrix_human_v_model_filtered_select2$identifier)


correlation_matrix_human_v_model_filtered_select2 <- correlation_matrix_human_v_model_filtered_select2 %>%
  dplyr::select(filename, identifier, model_type, classes, severity_rate_patch)

write.csv(correlation_matrix_human_v_model_filtered_select2, "../spreadsheets/human_vs_model_long_data_grid.csv")


correlation_matrix_human_v_model_filtered_select2 %>%
  count(filename, identifier, model_type) %>%
  filter(n > 1) %>%
  arrange(desc(n))

#View(correlation_matrix_human_v_model_filtered_select2)


correlation_matrix_human_v_model_filtered_select3 <- correlation_matrix_human_v_model_filtered_select2 %>%
  dplyr::filter(identifier %in% c("MW",
                                 "DG",
                                 "NW",
                                 "TB",
                                 "BR",
                                 "CC",
                                 "DG",
                                 #"Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5",
                                 #"Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.7_salth0.7",
                                 "Jan26_23-15-35_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5",
                                 "Jan26_23-15-35_2026_up0.95_dwn0.2_infg0.3_spth0.5_salth0.7",
                                 #"Feb03_16-24-20_2024_up0.8_dwn0.3_salth0.7",
                                 #"Feb14_15-53-04_2024_up0.8_dwn0.3_salth0.7",
                                 "Jan29_22-02-46_2026_up0.55_dwn0.15_infg0.3_spth0.65_salth0.5"
                                 ))

unique(correlation_matrix_human_v_model_filtered_select3$identifier)

library(data.table)

tmp <- correlation_matrix_human_v_model_filtered_select3 %>%
  mutate(
    severity_rate_patch = readr::parse_number(
      na_if(as.character(severity_rate_patch), "-")
    )
  )

#dt <- as.data.table(tmp)

#dt_nodups <- dt[, if (.N == 1) .SD, by = .(filename, identifier, model_type)]


# pivot the table
data_wide <- dcast(
  as.data.table(tmp),
  filename ~ identifier + model_type,
  value.var = "severity_rate_patch",
  fun.aggregate = mean,
  na.rm = TRUE
)

##data_wide <- dcast(
#  dt_nodups,
#  filename ~ identifier + model_type,
#  value.var = "severity_rate_patch"
#)

write.csv(data_wide, "../spreadsheets/human_vs_model_wide_data_grid.csv")

# drop columns that start with Feb01_ or Aug31_ *these are old model versions that we dont want to include in the analysis*
data_wide_na_rm <- data_wide %>%
  dplyr::select(-dplyr::matches("^(Feb01_|Aug31_|Feb03_04)")) %>%
  dplyr::select(-dplyr::matches("May11_00-29-56_2021")) # remove this specific column


data_wide_na_rm_reordered <- data_wide_na_rm %>%
  dplyr::select(
    -dplyr::ends_with("_computer_vision"),
    -dplyr::ends_with("_human"),
    dplyr::ends_with("_computer_vision"),
    dplyr::ends_with("_human")
  )


write.csv(data_wide_na_rm_reordered, "../spreadsheets/human_vs_model_wide_data_grid.csv")


# drop Dave since he didnt recieve great instructions
data_wide_na_rm_reordered <- data_wide_na_rm_reordered %>%
  dplyr::select(-dplyr::matches("DG_human"))
  

# Average out the human ratings
data_wide_na_rm_reordered_humans_averaged <- data_wide_na_rm_reordered %>%
  dplyr::mutate(
    mean_human_severity   = rowMeans(dplyr::select(., dplyr::matches("_human")), na.rm = TRUE),
    median_human_severity = apply(dplyr::select(., dplyr::matches("_human")), 1, median, na.rm = TRUE),
    sd_human_severity     = apply(dplyr::select(., dplyr::matches("_human")), 1, sd, na.rm = TRUE),
    n_humans              = rowSums(!is.na(dplyr::select(., dplyr::matches("_human"))))
  ) %>%
  # remove any NaN
  drop_na()

# rename BR_Human to Human_Rater1; CC_Human to Human_Rater2; MW_Human to Human_Rater3; NW_Human to Human_Rater4; TB_Human to Human_Rater5
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("BR_human", "Human_Rater1", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("CC_human", "Human_Rater2", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("MW_human", "Human_Rater3", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("NW_human", "Human_Rater4", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("TB_human", "Human_Rater5", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("mean_human_severity", "Mean_Human_Rating", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("median_human_severity", "Median_Human_Rating", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.7_salth0.7_computer_vision", "Jan22_10-21-00_2026_sal0.7", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Feb03_16-24-20_2024_up0.8_dwn0.3_salth0.7_computer_vision", "Feb03_16-24-20_2024", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Feb14_15-53-04_2024_up0.8_dwn0.3_salth0.7_computer_vision", "Feb14_15-53-04_2024", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Jan26_23-15-35_2026_up0.95_dwn0.2_infg0.3_spth0.5_salth0.7_computer_vision", "Jan26_23-15-35_2026_sal0.7", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Jan26_23-15-35_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5_computer_vision", "Jan26_23-15-35_2026_sal0.5", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5_computer_vision", "Jan22_10-21-00_2026_sal0.5", colnames(data_wide_na_rm_reordered_humans_averaged))
colnames(data_wide_na_rm_reordered_humans_averaged) <- gsub("Jan29_22-02-46_2026_up0.55_dwn0.15_infg0.3_spth0.65_salth0.5", "Jan29_22-02-46_2026_sal0.5", colnames(data_wide_na_rm_reordered_humans_averaged))


View(data_wide_na_rm_reordered_humans_averaged)

```

# ------------------------------

## Pairs plot with Pearson's Corr. Coeff.
Human v model correlation matrices. 
```{r, prep and viz, pearsons correlation, warning=FALSE}

  
  lower.panel <- function(x, y){
    points(x, y, pch = 19)
  
    r <- round(cor(x, y), 2)
    txt <- sprintf("R = %.2f", r)
  
    # limit to 17 characters (truncate if longer)
    txt <- substr(txt, 1, 17)
  
  
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
  
    x0 <- 0.5
    y0 <- 0.9
    off <- 0.01     # outline thickness
    cex_txt <- 1.5   # text size
    outline_col <- "black"
  
    # 8-direction outline offsets
    offsets <- matrix(c(
      -off,  0,
       off,  0,
       0,   -off,
       0,    off,
      -off, -off,
      -off,  off,
       off, -off,
       off,  off
    ), ncol = 2, byrow = TRUE)
  
    # draw outline
    for(i in 1:nrow(offsets)){
      text(x0 + offsets[i, 1], y0 + offsets[i, 2],
           txt, col = outline_col, font = 2, cex = cex_txt)
    }
  
    # draw main text on top
    text(x0, y0, txt, col = "#00E5FF", font = 2, cex = cex_txt)
  }
  
  
  df_plot <- data_wide_na_rm_reordered_humans_averaged %>%
    select(-n_humans, -sd_human_severity)   # keep filename, drop stats you don't want
  
  df_pairs <- df_plot %>%
    select(-filename) %>%
    select(where(is.numeric))
  
  # truncate column names to 17 characters
  colnames(df_pairs) <- substr(colnames(df_pairs), 1, 12)
  
  
  library(Cairo)
  
  CairoPDF("../figs_for_pub/correlation_plots/pairs_plot_human_vs_model_grid.pdf",
           width = 12, height = 12,
           family = "Arial")

labels_17 <- substr(colnames(df_pairs), 1, 12)

pairs(
  df_pairs,
  labels = labels_17,
  upper.panel = NULL,
  lower.panel = lower.panel,
  cex.labels = 1.4
)
  
  dev.off()


#pairs(data_wide_na_rm_reordered_humans_averaged[,c(6,25:37)], upper.panel = NULL, lower.panel = lower.panel)


#select_comp<-data_wide_na_rm_reordered_humans_averaged %>%
#  select(1,6,8,17)

#pairs(select_comp[,2:4], upper.panel = NULL, lower.panel = lower.panel)


```

## Print out top correlations (Human v patch severity model outputs)

```{r}

# 1) Get only the computer vision columns
cv_cols <- names(data_wide_na_rm_reordered_humans_averaged) %>%
  .[grepl("_202", .)]

cv_cols

# 2) Compute correlation vs mean_human_severity for each CV model
cv_correlations <- tibble(model = cv_cols) %>%
  mutate(
    correlation = sapply(model, function(col) {
      cor(
        data_wide_na_rm_reordered_humans_averaged[[col]],
        data_wide_na_rm_reordered_humans_averaged$Mean_Human_Rating,
        use = "pairwise.complete.obs"
      )
    })
  ) %>%
  arrange(desc(abs(correlation)))  # sort by strongest correlation (positive or negative)

# 3) Print top 5
cv_correlations %>% slice(1:50)



```

# Inter-rater reliability

```{r}

#### Human inter-rater reliability (ICC) ######
df <- data_wide_na_rm_reordered_humans_averaged %>%
  select(filename, starts_with("Human_Rater"), everything())

human_cols <- names(df) %>% str_subset("^Human_Rater\\d+$")

cv_cols <- df %>%
  select(-all_of(human_cols), -filename) %>%
  select(where(is.numeric)) %>%
  names() %>%
  .[. != "Mean_Human_Rating" & . != "Median_Human_Rating" & . != "sd_human_severity" & . != "n_humans"]

human_cols
cv_cols



# Calculate ICC for human raters
human_mat <- df %>% select(all_of(human_cols))

# ICC(2,1) and ICC(2,k) are common when raters are considered a random sample
icc_out <- psych::ICC(human_mat)

print(icc_out)

# Calculate ICC for CV models (not that informative tbh)
cv_mat <- df %>% select(all_of(cv_cols))
icc_out_cv <- psych::ICC(cv_mat)
print(icc_out_cv)
```



## Rater difference vs image difference


```{r}

##### rater difference vs image difference #####

df_long_humans <- df %>%
  pivot_longer(cols = all_of(human_cols), names_to = "rater", values_to = "human_score")

df_long_humans <- df_long_humans %>%
  dplyr::select(
    -all_of(c(
      #"Jan22_10-21-00_2026_sal0.5",
     # "Jan22_10-21-00_2026_sal0.7",
      "Jan26_23-15-35_2026_sal0.5",
      "Jan29_22-02-46_2026_sal0.5_computer_vision"
    ))
  )

m_human_vc <- lmer(human_score ~ 1 + (1|filename) + (1|rater), data = df_long_humans, REML = TRUE)
summary(m_human_vc)

# Variance components
vc <- as.data.frame(VarCorr(m_human_vc))
vc

var_image <- vc %>% filter(grp == "filename") %>% pull(vcov)
var_rater <- vc %>% filter(grp == "rater")    %>% pull(vcov)
var_resid <- attr(VarCorr(m_human_vc), "sc")^2
total_var <- var_image + var_rater + var_resid

tibble(
  component = c("image", "rater", "residual"),
  variance  = c(var_image, var_rater, var_resid),
  pct       = 100 * variance / total_var
)
```

So it looks like 84.8% of variance is due to differences between images, 2.83% is due to raters, and 12.3% is residual. 

## Concordance of each CV model with humans
```{r}
#### Agreement of each CV model with humans ####

str(df)

df_human_summary <- df %>%
  mutate(
    human_mean = rowMeans(across(all_of(human_cols)), na.rm = TRUE),
    human_sd   = apply(select(., all_of(human_cols)), 1, sd, na.rm = TRUE)
  )
```

### Functions you dont need to re-run (below)
Below is ccc function
```{r}
metric_vs_reference <- function(pred, ref) {
  ok <- is.finite(pred) & is.finite(ref)
  pred <- pred[ok]; ref <- ref[ok]

  ccc <- tryCatch(DescTools::CCC(pred, ref)$rho.c[1], error = function(e) NA_real_)

  tibble(
    n    = length(ref),
    CCC  = as.numeric(ccc),
    cor  = as.numeric(cor(pred, ref)),
    MAE  = mean(abs(pred - ref)),
    RMSE = sqrt(mean((pred - ref)^2)),
    bias = mean(pred - ref)
  )
}

```

## Bland-Altman, individual correlation plots, relationship between cv error and human error
```{r}

cv_vs_mean_tbl <- purrr::map_dfr(cv_cols, \(cv) {
  out <- metric_vs_reference(
    dplyr::pull(df_human_summary, all_of(cv)),
    df_human_summary$human_mean
  )
  out$model <- cv
  out
}) %>%
  dplyr::relocate(model)

cv_vs_mean_tbl


cv_vs_each_human <- expand_grid(model = cv_cols, rater = human_cols) %>%
  mutate(metrics = map2(model, rater, \(m, r) {
    metric_vs_reference(df[[m]], df[[r]])
  })) %>%
  unnest(metrics)

cv_vs_each_human

cv_vs_each_human2 <- cv_vs_each_human %>%
  mutate(
    CCC  = map_dbl(CCC,  ~ .x[1]),
    cor  = map_dbl(cor,  ~ .x[1]),
    MAE  = map_dbl(MAE,  ~ .x[1]),
    RMSE = map_dbl(RMSE, ~ .x[1]),
    bias = map_dbl(bias, ~ .x[1])
  )

cv_vs_each_human2 %>%
  dplyr::filter(!rater %in% c("Mean_Human_Rating", "Median_Human_Rating")) %>%
  group_by(model) %>%
  summarize(
    CCC_mean = mean(CCC, na.rm = TRUE),
    CCC_sd   = sd(CCC,  na.rm = TRUE),
    MAE_mean = mean(MAE, na.rm = TRUE),
    RMSE_mean= mean(RMSE, na.rm = TRUE),
    .groups = "drop"
  )


##### Plots: CV vs human mean + Bland–Altman ######

plot_scatter_with_cor <- function(cv_name, df) {

  d <- df %>%
    dplyr::select(Mean_Human_Rating, !!rlang::sym(cv_name)) %>%
    rename(cv = !!rlang::sym(cv_name)) %>%
    tidyr::drop_na()
  
  r <- suppressWarnings(cor(d$Mean_Human_Rating, d$cv))

  short_name <- ifelse(
    nchar(cv_name) > 19,
    paste0(substr(cv_name, 1, 19), ""),
    cv_name
  )

  ggplot(d, aes(x = Mean_Human_Rating, y = cv)) +
    geom_point(alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = expression(bold("A.")~Mean~Expert~Severity~Rating~vs.~Model),
      x = "Mean of Expert Disease Severity Rating",
      y = "Model Severity Rating",
      subtitle = paste0(
  short_name, " | r = ", sprintf("%.3f", r),
  " | n = ", nrow(d))
    ) +
    theme_classic()
}

wrap_plots(lapply(cv_cols, plot_scatter_with_cor, df = df))


```



```{r}
plot_ba_panel <- function(df, cv_name) {
  d <- df %>%
    transmute(
      mean_xy = (human_mean + .data[[cv_name]])/2,
      diff_xy = .data[[cv_name]] - human_mean
    ) %>%
    drop_na()

  bias <- mean(d$diff_xy)
  loa  <- bias + c(-1, 1) * 1.96 * sd(d$diff_xy)
  
  short_name <- ifelse(
  nchar(cv_name) > 19,
  paste0(substr(cv_name, 1, 19), ""),
  cv_name)

  ggplot(d, aes(mean_xy, diff_xy)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = bias) +
    geom_hline(yintercept = loa[1], linetype = 2) +
    geom_hline(yintercept = loa[2], linetype = 2) +
    labs(
      title = expression(bold("B.")~Bland~-~Altman),
      subtitle = paste0(
        short_name,
        "  |  Bias=", sprintf("%.2f", bias),
        "  |  LoA=[", sprintf("%.2f", loa[1]), ", ", sprintf("%.2f", loa[2]), "]"
      ),
      x = "Mean of (Expert Mean, Model) (%)",
      y = "Model − Expert Mean (%)"
    ) +
    theme_classic()
}

# helper to make safe filenames
safe_name <- function(x) str_replace_all(x, "[^A-Za-z0-9_\\-]+", "_")

save_model_agreement_plots <- function(df, cv_cols, out_dir = "agreement_plots",
                                       width = 10, height = 5, dpi = 300,
                                       also_pdf = TRUE) {
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  for (cv_name in cv_cols) {
    pA <- plot_scatter_with_cor(cv_name, df)
    pB <- plot_ba_panel(df, cv_name)

    p  <- pA + pB + plot_layout(ncol = 2)

    fn_base <- file.path(out_dir, paste0("human_vs_", safe_name(cv_name)))

    ggsave(paste0(fn_base, ".png"), plot = p, width = width, height = height, dpi = dpi)
    if (also_pdf) {
      ggsave(paste0(fn_base, ".pdf"), plot = p, width = width, height = height)
    }
  }

  invisible(TRUE)
}




p_ba <- wrap_plots(lapply(cv_cols, function(x) plot_ba_panel(df_human_summary, x)))
print(p_ba)

logit01 <- function(x, eps = 0.001) {
  # x in 0..100 -> p in (0,1)
  p <- pmin(pmax(x/100, eps), 1 - eps)
  qlogis(p)
}
inv_logit01 <- function(z) 100 * plogis(z)


fit_human_on_cv <- function(cv_name) {
  d <- df %>%
    select(filename, all_of(human_cols), all_of(cv_name)) %>%
    pivot_longer(all_of(human_cols), names_to = "rater", values_to = "human_score") %>%
    mutate(cv_score = .data[[cv_name]]) %>%
    drop_na(human_score, cv_score) %>%
    mutate(
      y = logit01(human_score),
      x = logit01(cv_score)
    )

  # random intercepts for image + rater
  lmer(y ~ x + (1|filename) + (1|rater), data = d, REML = TRUE)
}

models_human_on_cv <- setNames(lapply(cv_cols, fit_human_on_cv), cv_cols)

lapply(models_human_on_cv, summary)

##### Does CV struggle when humans disagree? #########

cv_error_vs_disagreement <- function(cv_name) {
df_human_summary %>%
  tidyr::pivot_longer(
    cols = all_of(cv_cols),
    names_to = "cv_name",
    values_to = "cv"
  ) %>%
  mutate(abs_err = abs(cv - human_mean)) %>%
  ggplot(aes(x = human_sd, y = abs_err)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~ cv_name) +
  labs(
    title = "Absolute CV error vs Human disagreement",
    x = "Human SD across 5 raters",
    y = "|CV − Human mean|"
  ) +
      theme_classic()
}


wrap_plots(lapply(cv_cols, function(x) plot_ba_panel(df_human_summary, x)))
save_model_agreement_plots(df_human_summary, cv_cols, out_dir = "agreement_plots")
list.files("agreement_plots")
browseURL(file.path(getwd(), "agreement_plots"))

cv_error_vs_disagreement(cv_name)


# quantile regression
# looks like error variance increases with rater disagreement
# worst errors are likely conditional on label ambiguity, not model instability
df_human_summary %>%
  tidyr::pivot_longer(
    cols = all_of(cv_cols),
    names_to = "cv_name",
    values_to = "cv"
  ) %>%
  mutate(abs_err = abs(cv - human_mean)) %>%
    ggplot(., aes(human_sd, abs_err)) +
    geom_point(alpha = 0.3) +
    geom_quantile(quantiles = c(0.25, 0.5, 0.75), color = "blue") +
    facet_wrap(~ cv_name) +
    labs(
      title = "Quantile structure of CV error vs human disagreement",
      x = "Human SD",
      y = "|CV − Human mean|"
      ) +
    theme_classic()

df_long <- df_human_summary %>%
  tidyr::pivot_longer(
    cols = all_of(cv_cols),
    names_to = "cv_name",
    values_to = "cv"
    ) %>%
  mutate(abs_err = abs(cv - human_mean))

# not sure a linear model works here
lm_fit <- lm(abs_err ~ human_sd, data = df_long)

ggplot(df_long, aes(human_sd, resid(lm_fit))) +
  geom_point(alpha = 0.4) +
  geom_smooth(
  method = "gam",
  formula = y ~ s(x, k = 5),
  se = TRUE)+
  theme_classic()

# does the model disagree more at times when humans disagree with each other? 
# x axis is human sd deviation (ie when humans agree more)
# y axis is the mean absolute difference between the model and the sd bin
df_long %>%
  mutate(sd_bin = cut(human_sd, breaks = 10)) %>%
  ggplot(aes(x = sd_bin, y = abs_err)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  facet_wrap(~ cv_name) +
  labs(
    x = "Expert SD (Binned)",
    y = "Mean |Model − Expert Mean|"
  ) +
  theme_classic()
                                        
disagreement_concordance <- function(cv_name_value,
                                     out_dir = "disagreement_concordance",
                                     width = 3, height = 4, dpi = 300,
                                     save_pdf = TRUE) {

  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  p <- df_long %>%
    dplyr::filter(.data$cv_name == cv_name_value) %>%
    dplyr::mutate(sd_bin = cut(.data$human_sd, breaks = 10)) %>%
    ggplot2::ggplot(ggplot2::aes(x = sd_bin, y = .data$abs_err)) +
    ggplot2::stat_summary(fun = mean, geom = "point", size = 2) +
    ggplot2::stat_summary(fun.data = ggplot2::mean_cl_boot, geom = "errorbar", width = 0.2) +
    ggplot2::labs(
      x = "Human Expert SD (binned)",
      y = "|Model − Human Expert Mean|"
    ) +
    ylim(0,30) +
    ggplot2::theme_classic() +
    ggplot2::theme(axis.text = ggplot2::element_text(angle = 45, hjust = 1))

  safe <- gsub("[^A-Za-z0-9_\\-]+", "_", cv_name_value)
  png_path <- file.path(out_dir, paste0("disagreement_concord_", safe, ".png"))
  ggplot2::ggsave(png_path, p, width = width, height = height, dpi = dpi)

  if (isTRUE(save_pdf)) {
    pdf_path <- file.path(out_dir, paste0("disagreement_concord_", safe, ".pdf"))
    ggplot2::ggsave(pdf_path, p, width = width, height = height, device = cairo_pdf)
  }

  return(p)
}

invisible(lapply(unique(df_long$cv_name), disagreement_concordance))

# For most bins, the model’s mean absolute deviation from the human expert mean is small relative to the level of inter-rater variability observed among human experts.

ggplot(df_long, aes(human_sd, abs_err)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
  theme_classic()


```

Absolute prediction error increased nonlinearly with human rater disagreement. When raters largely agreed, model error was small and tightly bounded. In contrast, large errors occurred primarily at high disagreement levels, indicating that model performance is constrained by annotation uncertainty rather than systematic prediction bias.

## Investigate leave-one-out human CCC
The problem with relying on any given human is that one human clearly cannot define ground truth (this feels like a philosophical statement, but it's true...). 

```{r}
ccc_safe <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 3) return(NA_real_)

  out <- DescTools::CCC(x[ok], y[ok])

  # DescTools::CCC sometimes returns a list/data frame; extract robustly
  rho <- out$rho.c
  if (is.null(rho) && is.data.frame(out)) {
    # fallback if CCC returns a data.frame with column rho.c
    rho <- out[["rho.c"]][1]
  }

  as.numeric(rho)[1]
}


loo_consensus_ccc_one_model <- function(df, id_col, rater_cols, model_col) {
  id_col <- rlang::ensym(id_col)

  folds <- purrr::map(rater_cols, function(withheld) {
    others <- setdiff(rater_cols, withheld)

    df_fold <- df %>%
      dplyr::transmute(
        !!id_col,
        withheld = withheld,
        withheld_human = .data[[withheld]],
        consensus = rowMeans(dplyr::across(dplyr::all_of(others)), na.rm = TRUE),
        model = .data[[model_col]]
      )

    pairwise_cccs <- combn(others, 2, simplify = FALSE) %>%
      purrr::map_dbl(~ ccc_safe(df[[.x[1]]], df[[.x[2]]]))

    tibble::tibble(
      withheld = withheld,
      human_ccc = mean(pairwise_cccs, na.rm = TRUE),
      model_ccc = ccc_safe(df_fold$model, df_fold$consensus)
    )
  })

  dplyr::bind_rows(folds) %>%
    tidyr::pivot_longer(
      cols = c(human_ccc, model_ccc),
      names_to = "comparison",
      values_to = "ccc"
    ) %>%
    dplyr::mutate(
      comparison = dplyr::recode(
        comparison,
        human_ccc = "human_vs_consensus",
        model_ccc = "model_vs_consensus"
      ),
      model_name = model_col
    )
}

plot_and_save_loo_one_model <- function(df, id_col, rater_cols, model_col,
                                        out_dir = "loo_ccc_plots",
                                        width = 10, height = 4, dpi = 300,
                                        save_pdf = FALSE) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  loo_tbl <- loo_consensus_ccc_one_model(
    df = df, id_col = {{id_col}},
    rater_cols = rater_cols,
    model_col = model_col
  )

  human_baseline <- loo_tbl %>%
    dplyr::filter(comparison == "human_vs_consensus") %>%
    dplyr::summarise(baseline = mean(ccc, na.rm = TRUE)) %>%
    dplyr::pull(baseline)

  plot_df <- loo_tbl %>%
    dplyr::mutate(
      series = dplyr::if_else(
        comparison == "human_vs_consensus",
        "Human",
        model_col
      )
    )

  p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = withheld, y = ccc, color = series)) +
    ggplot2::geom_point(size = 2, alpha = 0.9) +
    ggplot2::geom_hline(yintercept = human_baseline, linetype = 2) +
    ggplot2::facet_wrap(~ comparison, scales = "free_y") +
    ggplot2::labs(
      title = paste0("LOO CCC: ", model_col),
      subtitle = paste0("Dashed line = mean peer CCC (", sprintf("%.3f", human_baseline), ")"),
      x = "Withheld rater",
      y = "CCC"
    ) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "bottom")

  safe <- gsub("[^A-Za-z0-9_\\-]+", "_", model_col)
  png_path <- file.path(out_dir, paste0("loo_ccc_", safe, ".png"))

  ggplot2::ggsave(png_path, p, width = width, height = height, dpi = dpi)

  if (isTRUE(save_pdf)) {
    pdf_path <- file.path(out_dir, paste0("loo_ccc_", safe, ".pdf"))
    ggplot2::ggsave(pdf_path, p, width = width, height = height, device = grDevices::cairo_pdf)
  }

  invisible(list(plot = p, data = loo_tbl, png = png_path))
}

rater_cols <- c("Human_Rater1", "Human_Rater2", "Human_Rater3", "Human_Rater4", "Human_Rater5")

results <- purrr::map(
  cv_cols,
  ~ plot_and_save_loo_one_model(
    df = df_human_summary,
    id_col = filename,
    rater_cols = rater_cols,
    model_col = .x,
    out_dir = "loo_ccc_plots",
    save_pdf = FALSE
  )
)

purrr::map_chr(results, "png")

plot_and_save_loo_dumbbell_one_model <- function(df, id_col, rater_cols, model_col,
                                                 out_dir = "loo_ccc_dumbbell",
                                                 width = 4.5, height = 3.2, dpi = 300,
                                                 save_pdf = TRUE) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  loo_tbl <- loo_consensus_ccc_one_model(
    df = df, id_col = {{id_col}},
    rater_cols = rater_cols,
    model_col = model_col
  )

  wide_tbl <- loo_tbl %>%
    dplyr::select(withheld, comparison, ccc) %>%
    tidyr::pivot_wider(names_from = comparison, values_from = ccc) %>%
    dplyr::rename(
      human_ccc = human_vs_consensus,
      model_ccc = model_vs_consensus
    ) %>%
    dplyr::mutate(
      delta_ccc = model_ccc - human_ccc,
      withheld_lab = gsub("^Human_Rater", "R", withheld)
    )

  human_baseline <- mean(wide_tbl$human_ccc, na.rm = TRUE)

  out_tbl <- wide_tbl %>%
    dplyr::mutate(
      model_name = model_col,
      human_baseline_mean = human_baseline
    ) %>%
    dplyr::select(model_name, withheld, withheld_lab, human_ccc, model_ccc, delta_ccc, human_baseline_mean)

  short_name <- ifelse(
    nchar(model_col) > 19,
    paste0(substr(model_col, 1, 19), ""),
    model_col
  )

  # long format for colored endpoints
  endpoints <- out_tbl %>%
    tidyr::pivot_longer(cols = c(human_ccc, model_ccc),
                        names_to = "endpoint", values_to = "ccc") %>%
    dplyr::mutate(endpoint = dplyr::recode(endpoint,
                                           human_ccc = "Human (peer CCC)",
                                           model_ccc = "Model"))

  p <- ggplot2::ggplot(out_tbl, ggplot2::aes(y = withheld_lab)) +
    ggplot2::geom_segment(
      ggplot2::aes(x = human_ccc, xend = model_ccc, yend = withheld_lab),
      linewidth = 0.8, color = "grey40", alpha = 0.7
    ) +
    ggplot2::geom_point(data = endpoints,
                        ggplot2::aes(x = ccc, color = endpoint),
                        size = 2.6) +
    ggplot2::geom_vline(xintercept = human_baseline, linetype = 2) +
    ggplot2::scale_color_manual(values = c("Human (peer CCC)" = "salmon",
                                           "Model" = "steelblue")) +
    ggplot2::coord_cartesian(xlim = c(0.8, 1.0)) +
    ggplot2::labs(
      title = paste0("LOO CCC vs consensus: ", short_name),
      x = "Concordance correlation coefficient",
      y = "Withheld rater",
      color = NULL
    ) +
    ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "bottom")

  safe <- gsub("[^A-Za-z0-9_\\-]+", "_", model_col)
  png_path <- file.path(out_dir, paste0("loo_dumbbell_", safe, ".png"))
  csv_path <- file.path(out_dir, paste0("loo_table_", safe, ".csv"))

  ggplot2::ggsave(png_path, p, width = width, height = height, dpi = dpi)

  if (isTRUE(save_pdf)) {
    pdf_path <- file.path(out_dir, paste0("loo_dumbbell_", safe, ".pdf"))
    ggplot2::ggsave(pdf_path, p, width = width, height = height, device = grDevices::cairo_pdf)
  } else {
    pdf_path <- NA_character_
  }

  utils::write.csv(out_tbl, csv_path, row.names = FALSE)

  invisible(list(plot = p, loo_long = loo_tbl, loo_table = out_tbl,
                 png = png_path, pdf = pdf_path, csv = csv_path))
}

results <- purrr::map(
  cv_cols,
  ~ plot_and_save_loo_dumbbell_one_model(
    df = df_human_summary,
    id_col = filename,
    rater_cols = rater_cols,
    model_col = .x,
    out_dir = "loo_ccc_dumbbell",
    save_pdf = FALSE
  )
)

purrr::map_chr(results, "png")


```
Model–consensus concordance was consistently high and largely invariant to which rater was withheld, whereas human–consensus concordance varied substantially depending on the rater.


Note human_baseline_mean is the mean leave-one-out human–consensus concordance.

## Dumbell plots of LOO vs Model-consensus

```{r}
plot_and_save_loo_dumbbell_one_model <- function(df, id_col, rater_cols, model_col,
                                                 out_dir = "loo_ccc_dumbbell",
                                                 width = 4.5, height = 3.2, dpi = 300,
                                                 save_pdf = TRUE) {

  # deps: dplyr, tidyr, purrr, ggplot2, rlang, DescTools
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

  # LOO CCC long table (your existing helper)
  loo_tbl <- loo_consensus_ccc_one_model(
    df = df, id_col = {{id_col}},
    rater_cols = rater_cols,
    model_col = model_col
  )

  # Wide table: one row per withheld rater
  wide_tbl <- loo_tbl %>%
    dplyr::select(withheld, comparison, ccc) %>%
    tidyr::pivot_wider(names_from = comparison, values_from = ccc) %>%
    dplyr::rename(
      human_ccc = human_vs_consensus,
      model_ccc = model_vs_consensus
    ) %>%
    dplyr::mutate(
      delta_ccc = model_ccc - human_ccc
    )

  # Baseline (mean human CCC)
  human_baseline <- wide_tbl %>%
    dplyr::summarise(b = mean(human_ccc, na.rm = TRUE)) %>%
    dplyr::pull(b)

  # Add baseline + model name for the CSV (handy for later merges)
  out_tbl <- wide_tbl %>%
    dplyr::mutate(
      model_name = model_col,
      human_baseline_mean = human_baseline
    ) %>%
    dplyr::select(model_name, withheld, human_ccc, model_ccc, delta_ccc, human_baseline_mean)
  
  short_name <- ifelse(
    nchar(model_col) > 19,
    paste0(substr(model_col, 1, 19), ""),
    model_col
  )

  # Plot (dumbbell)
  p <- ggplot2::ggplot(out_tbl, ggplot2::aes(y = withheld)) +
    ggplot2::geom_segment(
      ggplot2::aes(x = human_ccc, xend = model_ccc, yend = withheld),
      linewidth = 0.8, alpha = 0.7
    ) +
    ggplot2::geom_point(ggplot2::aes(x = human_ccc), size = 2.4) +
    ggplot2::geom_point(ggplot2::aes(x = model_ccc), size = 2.4) +
    ggplot2::geom_vline(xintercept = human_baseline, linetype = 2) +
    ggplot2::labs(
      title = paste0("LOO CCC vs consensus: ", short_name),
      #subtitle = paste0(
      #  "Per withheld rater: Human vs consensus (point) and Model vs consensus (point). ",
      #  "Dashed line = mean human CCC (", sprintf("%.3f", human_baseline), ")."
      #),
      x = "Concordance correlation coefficient",
      y = "Withheld rater"
    ) +
    xlim(0.8,1.0)+
    ggplot2::theme_classic()

  # Safe filenames
  safe <- gsub("[^A-Za-z0-9_\\-]+", "_", model_col)
  png_path <- file.path(out_dir, paste0("loo_dumbbell_", safe, ".png"))
  csv_path <- file.path(out_dir, paste0("loo_table_", safe, ".csv"))

  # Save outputs
  ggplot2::ggsave(png_path, p, width = width, height = height, dpi = dpi)

  if (isTRUE(save_pdf)) {
    pdf_path <- file.path(out_dir, paste0("loo_dumbbell_", safe, ".pdf"))
    ggplot2::ggsave(pdf_path, p, width = width, height = height, device = grDevices::cairo_pdf)
  } else {
    pdf_path <- NA_character_
  }

  utils::write.csv(out_tbl, csv_path, row.names = FALSE)

  invisible(list(plot = p, loo_long = loo_tbl, loo_table = out_tbl,
                 png = png_path, pdf = pdf_path, csv = csv_path))
}

rater_cols <- c("Human_Rater1", "Human_Rater2", "Human_Rater3", "Human_Rater4", "Human_Rater5")

results <- purrr::map(
  cv_cols,
  ~ plot_and_save_loo_dumbbell_one_model(
    df = df_human_summary,
    id_col = filename,
    rater_cols = rater_cols,
    model_col = .x,
    out_dir = "loo_ccc_dumbbell",
    save_pdf = TRUE
  )
)

# paths:
png_paths <- purrr::map_chr(results, "png")
csv_paths <- purrr::map_chr(results, "csv")

```
Figure X. LOO concordance of model and human raters with human consensus.
Dumbbells show CCC values for each withheld human rater, comparing agreement of the withheld rater and the CV model with the consensus of the remaining raters. The dashed line denotes the mean leave-one-out human–consensus CCC. Across all withheld raters, model concordance matches or exceeds that of individual human raters.

## Make a nice multi-panel figure
```{r}
# ---------- Theme ----------
theme_pub <- function(base_size = 11) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_markdown(face = "bold"),
      plot.subtitle = element_text(size = rel(0.95)),
      axis.text = element_text(color = "black"),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "none"
    )
}

panel_title <- function(letter, title) {
  paste0("<span style='font-weight:700;'>", letter, "</span> ", title)
}

make_panel_A <- function(df, model_col) {

  d <- df %>%
    transmute(
      x = human_mean,
      y = .data[[model_col]]
    ) %>%
    filter(is.finite(x), is.finite(y))

  r_val   <- cor(d$x, d$y, use = "complete.obs", method = "pearson")
  ccc_val <- ccc_safe(d$y, d$x)

  ann_text <- paste0(
    "r = ", sprintf("%.3f", r_val), "\n",
    "CCC = ", sprintf("%.3f", ccc_val)
  )

  # Bin x and estimate conditional quantiles
  q_df <- d %>%
    mutate(x_bin = cut(x, breaks = seq(0, 100, by = 5), include.lowest = TRUE)) %>%
    group_by(x_bin) %>%
    summarise(
      x_mid = mean(x, na.rm = TRUE),
      y_med = median(y, na.rm = TRUE),
      y_lo  = quantile(y, 0.10, na.rm = TRUE),
      y_hi  = quantile(y, 0.90, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(n >= 3)

  # Smooth the envelope
  q_grid <- tibble(x_mid = seq(min(q_df$x_mid), max(q_df$x_mid), length.out = 200))

  q_grid$y_lo  <- predict(loess(y_lo  ~ x_mid, data = q_df, span = 0.6), newdata = q_grid)
  q_grid$y_med <- predict(loess(y_med ~ x_mid, data = q_df, span = 0.6), newdata = q_grid)
  q_grid$y_hi  <- predict(loess(y_hi  ~ x_mid, data = q_df, span = 0.6), newdata = q_grid)

  q_grid <- q_grid %>%
    mutate(
      y_lo = pmax(y_lo, 0),
      y_hi = pmin(y_hi, 100)
    )

  ggplot(d, aes(x = x, y = y)) +
    geom_ribbon(
      data = q_grid,
      aes(x = x_mid, ymin = y_lo, ymax = y_hi),
      inherit.aes = FALSE,
      alpha = 0.18
    ) +
    geom_line(
      data = q_grid,
      aes(x = x_mid, y = y_med),
      inherit.aes = FALSE,
      linewidth = 1
    ) +
    geom_point(alpha = 0.35, size = 1.4) +
    geom_abline(intercept = 0, slope = 1, linetype = 2) +
    coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
    annotate(
      "text",
      x = 0, y = 100,
      label = ann_text,
      hjust = 0, vjust = 1.1,
      size = 4
    ) +
    labs(
      title = panel_title("A.", "Model vs human mean"),
      x = "Human mean severity (%)",
      y = "Model severity (%)"
    ) +
    theme_pub()
}

make_panel_B <- function(df, model_col) {
  d <- df %>%
    mutate(
      mean_pair = (human_mean + .data[[model_col]]) / 2,
      diff = .data[[model_col]] - human_mean
    )

  md <- mean(d$diff, na.rm = TRUE)
  sdv <- sd(d$diff, na.rm = TRUE)
  loa_u <- md + 1.96 * sdv
  loa_l <- md - 1.96 * sdv

  # annotation string
  ann <- paste0(
    " Bias = ", sprintf("%.2f", md), "%\n",
    "LoA = [", sprintf("%.2f", loa_l), ", ", sprintf("%.2f", loa_u), "]%"
  )

  ggplot(d, aes(x = mean_pair, y = diff)) +
    geom_point(alpha = 0.35) +
    geom_hline(yintercept = md) +
    geom_hline(yintercept = loa_u, linetype = 2) +
    geom_hline(yintercept = loa_l, linetype = 2) +
    # Put the annotation in the top-left of the panel
    annotate(
      "text",
      x = -Inf, y = Inf,
      label = ann,
      hjust = -0.15, vjust = 1.1,
      size = 4
    ) +
    labs(
      title = panel_title("B.", "Bland–Altman agreement"),
      x = "Mean of model and human mean (%)",
      y = "Model − human mean (%)"
      ) +
    theme_pub()
}


make_panel_C <- function(df, model_col, id_col, rater_cols,
                         shorten_raters = TRUE,
                         order_by = c("human", "delta"),
                         ccc_xlim = c(0.80, 1.00)) {

  order_by <- match.arg(order_by)

  loo_tbl <- loo_consensus_ccc_one_model(df, {{id_col}}, rater_cols, model_col)

  out_tbl <- loo_tbl %>%
    select(withheld, comparison, ccc) %>%
    pivot_wider(names_from = comparison, values_from = ccc) %>%
    rename(
      human_ccc = human_vs_consensus,
      model_ccc = model_vs_consensus
    ) %>%
    mutate(
      delta_ccc = model_ccc - human_ccc,
      withheld_lab = if (shorten_raters) gsub("^Human_Rater", "R", withheld) else withheld
    ) %>%
    mutate(withheld_lab = {
      ord <- if (order_by == "human") human_ccc else delta_ccc
      reorder(withheld_lab, ord)
    })

  human_baseline <- mean(out_tbl$human_ccc, na.rm = TRUE)

  # long format for colored endpoints
  endpoints <- out_tbl %>%
    select(withheld_lab, human_ccc, model_ccc) %>%
    pivot_longer(cols = c(human_ccc, model_ccc),
                 names_to = "endpoint", values_to = "ccc") %>%
    mutate(endpoint = recode(endpoint,
                             human_ccc = "Human (peer CCC)",
                             model_ccc = "Model"))

  ggplot(out_tbl, aes(y = withheld_lab)) +
    geom_segment(aes(x = human_ccc, xend = model_ccc, yend = withheld_lab),
                 linewidth = 0.9, color = "black", alpha = 0.7) +
    geom_point(data = endpoints,
               aes(x = ccc, color = endpoint),
               size = 2.6) +
    geom_vline(xintercept = human_baseline, linetype = 2) +
    coord_cartesian(xlim = ccc_xlim) +
    labs(
      title = panel_title("C.", "LOO concordance vs consensus"),
      #subtitle = paste0("Dashed = mean LOO human CCC (", sprintf("%.3f", human_baseline), ")"),
      x = "CCC",
      y = "Withheld rater"
    ) +
    theme_pub() +
    theme(
      legend.position = c(0.02, 0.98),     # inside, upper-left
      legend.justification = c(0, 1),      # anchor legend's top-left corner
      legend.background = element_rect(fill = scales::alpha("white", 0.8), color = NA),
    legend.title = element_blank())
}



make_panel_D <- function(df, model_col, add_ref = FALSE, ref_y = 5) {
  d <- df %>%
    mutate(abs_err = abs(.data[[model_col]] - human_mean))

  p <- ggplot(d, aes(x = human_sd, y = abs_err)) +
    geom_point(alpha = 0.30) +
    geom_rug(sides = "b", alpha = 0.15) +
    geom_smooth(method = "gam", formula = y ~ s(x, k = 5), se = TRUE) +
    labs(
      title = panel_title("D.", "Model error tracks rater disagreement"),
      x = "Human SD across raters",
      y = "|Model − human mean| (%)"
    ) +
    theme_pub()

  if (isTRUE(add_ref)) {
    p <- p + geom_hline(yintercept = ref_y, linetype = 3, alpha = 0.6)
  }

  p
}


```


```{r}
library(quantreg)

str(df_human_summary)

best_model <- "Jan26_23-15-35_2026_sal0.5"
rater_cols <- c("Human_Rater1","Human_Rater2","Human_Rater3","Human_Rater4","Human_Rater5")

pC <- make_panel_C(df_human_summary, best_model, filename, rater_cols, order_by = "human")
pB <- make_panel_B(df_human_summary, best_model)
pD <- make_panel_D(df_human_summary, best_model)



make_4panel_figure <- function(df, model_col, id_col, rater_cols,
                               sev_xlim = c(0, 100),
                               sev_ylim = c(0, 100),
                               err_ylim = NULL,
                               loo_order_by = c("human", "delta"),
                               shorten_raters = TRUE,
                               add_ref_err = FALSE,
                               ref_err_y = 5) {

  loo_order_by <- match.arg(loo_order_by)

  # If you don’t pass err_ylim, compute it from data (robust-ish)
  if (is.null(err_ylim)) {
    err <- abs(df[[model_col]] - df$human_mean)
    err_ylim <- c(0, max(err, na.rm = TRUE))
  }

  pA <- make_panel_A(df, model_col) +
    coord_equal(xlim = sev_xlim, ylim = sev_ylim)

  pB <- make_panel_B(df, model_col)   # uses annotated version you updated

  pC <- make_panel_C(
    df, model_col, {{id_col}}, rater_cols,
    shorten_raters = shorten_raters,
    order_by = loo_order_by
  )

  pD <- make_panel_D(df, model_col, add_ref = add_ref_err, ref_y = ref_err_y) +
    coord_cartesian(ylim = err_ylim)

top    <- pA | pB
bottom <- pC | pD

(top / bottom) +
  patchwork::plot_layout(
    widths  = c(1.5, 1),   # left col wider so A/C look balanced
    heights = c(2, 1)    # top row taller
  )

}



fig4 <- make_4panel_figure(
  df = df_human_summary,
  model_col = best_model,
  id_col = filename,
  rater_cols = rater_cols,
  sev_xlim = c(0, 100),
  sev_ylim = c(0, 100),
  loo_order_by = "human",      # or "delta"
  shorten_raters = TRUE,
  add_ref_err = FALSE          # set TRUE if you want a horizontal error threshold line
)

fig4
ggsave("FigureX_model_validation_4panel.png", fig4, width = 10, height = 8, dpi = 300)
ggsave("FigureX_model_validation_4panel.pdf", fig4, width = 10, height = 8, device = cairo_pdf)


```

# -----------------------------

# Repeat above, but for saliency methods 


next chunk below are helper functions -- run once and then you dont have to run them again if you're in the same R session.

```{r}

# FULL PIPELINE WRAPPED INTO FUNCTIONS
# - loops over saliency methods: Gradient_sr2, Gradient-SG_sr2, GradCAM_sr2, DeepLift_sr2
# - runs: combine/clean -> wide -> averaged humans -> pairs plot PDFinstall.packages("xfun")

# - runs: CV-vs-human correlations, ICC (humans + CV), variance components (lmer),
#         CV agreement metrics vs human mean, CV vs each human,
#         scatter plots + Bland–Altman plots,
#         mixed model of human-on-CV (logit), error vs disagreement plots
#
# Assumes these objects already exist in your workspace:
#   correlation_matrix_human_v_model
#   concatenated_df_trimmed




metric_vs_reference <- function(pred, ref) {
  ok <- is.finite(pred) & is.finite(ref)
  pred <- pred[ok]; ref <- ref[ok]

  ccc <- tryCatch(DescTools::CCC(pred, ref)$rho.c[1], error = function(e) NA_real_)

  tibble(
    n    = length(ref),
    CCC  = as.numeric(ccc),
    cor  = as.numeric(cor(pred, ref)),
    MAE  = mean(abs(pred - ref)),
    RMSE = sqrt(mean((pred - ref)^2)),
    bias = mean(pred - ref)
  )
}

logit01 <- function(x, eps = 0.001) {
  p <- pmin(pmax(x/100, eps), 1 - eps)
  qlogis(p)
}

lower.panel <- function(x, y){
  points(x, y, pch = 19)

  r <- round(cor(x, y, use = "pairwise.complete.obs"), 2)
  txt <- paste0("R = ", r)

  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))

  x0 <- 0.5
  y0 <- 0.9
  off <- 0.01
  cex_txt <- 1.5
  outline_col <- "black"

  offsets <- matrix(c(
    -off,  0,
     off,  0,
     0,   -off,
     0,    off,
    -off, -off,
    -off,  off,
     off, -off,
     off,  off
  ), ncol = 2, byrow = TRUE)

  for(i in 1:nrow(offsets)){
    text(x0 + offsets[i, 1], y0 + offsets[i, 2],
         txt, col = outline_col, font = 2, cex = cex_txt)
  }
  text(x0, y0, txt, col = "#00E5FF", font = 2, cex = cex_txt)
}

plot_cv_scatter <- function(df_human_summary, cv_name) {
  ggplot(df_human_summary, aes(x = human_mean, y = .data[[cv_name]])) +
    geom_point(alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1) +
    labs(
      title = paste("CV vs Human Mean:", cv_name),
      x = "Human mean severity (%)",
      y = "CV severity (%)"
    ) +
    coord_equal()
}

plot_ba <- function(df_human_summary, cv_name) {
  d <- df_human_summary %>%
    transmute(
      mean_xy = (human_mean + .data[[cv_name]])/2,
      diff_xy = .data[[cv_name]] - human_mean
    ) %>%
    drop_na()

  bias <- mean(d$diff_xy)
  loa  <- bias + c(-1, 1) * 1.96 * sd(d$diff_xy)

  ggplot(d, aes(mean_xy, diff_xy)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = bias) +
    geom_hline(yintercept = loa[1], linetype = 2) +
    geom_hline(yintercept = loa[2], linetype = 2) +
    labs(
      title = paste("Bland–Altman:", cv_name),
      x = "Mean of (Expert Mean, Model) (%)",
      y = "Model − Expert Mean (%)"
    )
}

cv_error_vs_disagreement <- function(df_human_summary, cv_name) {
  df_human_summary %>%
    mutate(abs_err = abs(.data[[cv_name]] - human_mean)) %>%
    ggplot(aes(x = human_sd, y = abs_err)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = TRUE) +
    labs(
      title = paste("Absolute Model Error vs Expert Disagreement:", cv_name),
      x = "SD Across Experts",
      y = "|Model − Expert Mean|"
    )
}


fit_human_on_cv <- function(df, human_cols, cv_name) {
  d <- df %>%
    select(filename, all_of(human_cols), all_of(cv_name)) %>%
    pivot_longer(all_of(human_cols), names_to = "rater", values_to = "human_score") %>%
    mutate(cv_score = .data[[cv_name]]) %>%
    drop_na(human_score, cv_score) %>%
    mutate(
      y = logit01(human_score),
      x = logit01(cv_score)
    )
  print("################## fit human on cv function ####################")
  print("")

  lmer(y ~ x + (1|filename) + (1|rater), data = d, REML = TRUE)
}
```

## You may have to edit model prefixes in chunk below... otherwise dont edit it... it's painful 

There are two edits that needs to be made to make sure you catch the model identifier:
*`stringr::str_detect(identifier, "^(Jan22_|Jan26_|Jan29_)")`
*`  keep_prefix = c(`

Eventually I'll make this more robust, but this works for now. 

```{r}

inv_logit01 <- function(z) 100 * plogis(z)


make_human_vs_model <- function(
  correlation_matrix_human_v_model,
  concatenated_df_trimmed,
  saliency_col,
  keep_humans = c("MW","DG","NW","TB","BR","CC"),
  keep_prefix = c("Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.7","Jan26_23-15-35_2026_up0.95_dwn0.2_infg0.3_spth0.5","Jan29_22-02-46_2026_up0.55__dwn0.15_infg0.3_spth0.65"),
  drop_models_regex = "^(Feb01_|Aug31_|Feb03_04)|May11_00-29-56_2021",
  out_prefix = "human_vs_model",
  out_dir = "."
) {

  # 1) bind + force consistent types
  combined <- dplyr::bind_rows(
    correlation_matrix_human_v_model %>% dplyr::mutate(dplyr::across(dplyr::everything(), as.character)),
    concatenated_df_trimmed         %>% dplyr::mutate(dplyr::across(dplyr::everything(), as.character))
  )

  # 2) resolve analysis column robustly

  # scalar / non-saliency metrics that should NEVER get _sr2
  scalar_metrics <- c(
    "severity_rate_patch",
    "severity_patch_rate"
  )
  
  if (saliency_col %in% names(combined)) {
    # exact match always wins
    sal_col <- saliency_col
  
  } else if (saliency_col %in% scalar_metrics) {
    stop(
      "Scalar metric '", saliency_col,
      "' not found in combined data.\nAvailable columns:\n",
      paste(names(combined), collapse = ", ")
    )
  
  } else {
    # saliency-style fallback
    candidates <- c(
      saliency_col,
      paste0(saliency_col, "_sr2")
    )
    sal_col <- candidates[candidates %in% names(combined)][1]
  }
  
  if (is.na(sal_col)) {
    stop(
      "No matching analysis column found.\n",
      "Tried: ", paste(candidates, collapse = ", "), "\n",
      "Available columns:\n",
      paste(names(combined), collapse = ", ")
    )
  }
  
  cat("Starting with:", sal_col, "\n")



  # 3) human override: copy severity_rate_patch into the selected saliency column for human rows
  combined <- combined %>%
    dplyr::mutate(
      "{sal_col}" := dplyr::if_else(model_type == "human", severity_rate_patch, .data[[sal_col]])
    ) %>%
    dplyr::select(-severity_rate_patch)

  # 4) keep relevant cols + non-missing
  df0 <- combined %>%
    dplyr::select(filename, identifier, model_type, tray, classes, up_th, down_th, spor_th, inf_gate, dplyr::all_of(sal_col)) %>%
    dplyr::mutate(
      # turn "-", "", "NA" into NA then parse
      "{sal_col}" := readr::parse_number(dplyr::na_if(as.character(.data[[sal_col]]), "-")),
      "{sal_col}" := dplyr::na_if(.data[[sal_col]], NaN)
    ) %>%
    dplyr::filter(is.finite(.data[[sal_col]]))

  # 5) filename suffix logic
  df0 <- df0 %>%
    dplyr::mutate(filename = dplyr::case_when(
      grepl("5dpi", filename)  ~ filename,
      grepl("10dpi", filename) ~ filename,
      TRUE ~ paste0(filename, "_", tray)
    ))

  # 6) append threshold values into identifier if numeric (keeps separation by thresholds)
  df1 <- df0 %>%
    dplyr::mutate(identifier_base = identifier) %>%
    dplyr::mutate(
      up_th_num = suppressWarnings(as.numeric(up_th)),
      identifier = dplyr::if_else(
        !is.na(up_th_num) & !stringr::str_detect(identifier, "_up[0-9]"),
        paste0(identifier, "_up", up_th),
        identifier
      ),
  
      down_th_num = suppressWarnings(as.numeric(down_th)),
      identifier  = dplyr::if_else(
        !is.na(down_th_num) & !stringr::str_detect(identifier, "_dwn[0-9]"),
        paste0(identifier, "_dwn", down_th),
        identifier
      ),
  
      inf_gate_num = suppressWarnings(as.numeric(inf_gate)),
      identifier   = dplyr::if_else(
        !is.na(inf_gate_num) & !stringr::str_detect(identifier, "_infg[0-9]"),
        paste0(identifier, "_infg", inf_gate),
        identifier
      ),
  
      spor_th_num = suppressWarnings(as.numeric(spor_th)),
      identifier  = dplyr::if_else(
        !is.na(spor_th_num) & !stringr::str_detect(identifier, "_spth[0-9]"),
        paste0(identifier, "_spth", spor_th),
        identifier
      )
    ) %>%
    dplyr::select(-up_th_num, -down_th_num, -inf_gate_num, -spor_th_num)

  # 7) filter identifiers robustly:
  # - keep exact humans
  # - keep anything starting with Jan22 or Jan26 (so threshold-appended variants are kept)
    df2 <- tryCatch(
    {
      tmp <- df1 %>%
        dplyr::select(filename, identifier, model_type, classes, dplyr::all_of(sal_col)) %>%
        dplyr::filter(
          identifier %in% keep_humans |
            stringr::str_detect(identifier, "^(Jan22_|Jan26_|Jan29_)")
        ) %>%
        dplyr::mutate(
          "{sal_col}" := readr::parse_number(dplyr::na_if(as.character(.data[[sal_col]]), "-"))
        )
        tmp
    },
    error = function(e) {
      message("❌ df2 creation failed for ", sal_col, ": ", conditionMessage(e))
  
      message("Columns in df1: ", paste(names(df1), collapse = ", "))
      message("sal_col = ", sal_col)
      if (!("identifier" %in% names(df1))) message("MISSING: identifier")
      if (!("model_type" %in% names(df1))) message("MISSING: model_type")
      if (!(sal_col %in% names(df1))) message("MISSING: ", sal_col)
  
      NULL
    }
  )
  
  if (is.null(df2)) stop("Stopping because df2 failed (see error above).")

  # 8) wide pivot
  data_wide <- data.table::dcast(
    data.table::as.data.table(df2),
    filename ~ identifier + model_type,
    value.var = sal_col,
    fun.aggregate = mean,
    na.rm = TRUE
  )
  
  min_n <- 50
  keep_cols <- names(data_wide)[
  c(TRUE, colSums(!is.na(data_wide[, -1, with = FALSE])) >= min_n)]
  data_wide_pruned <- data_wide[, ..keep_cols]
  
  # remove any remaining rows with missing data
  data_wide <- data_wide_pruned %>%
    tidyr::drop_na()

  # 9) save grid
  sal_tag <- tolower(gsub("_sr2$", "", sal_col))
  out_grid <- file.path(out_dir, sprintf("%s_wide_data_grid_%s.csv", out_prefix, sal_tag))
  utils::write.csv2(data_wide, out_grid)

  # 10) drop old models + reorder
  data_wide_na_rm <- tibble::as_tibble(data_wide) %>%
    dplyr::select(-dplyr::matches(drop_models_regex))

  data_wide_na_rm_reordered <- data_wide_na_rm %>%
    dplyr::select(
      -dplyr::ends_with("_computer_vision"),
      -dplyr::ends_with("_human"),
      dplyr::ends_with("_computer_vision"),
      dplyr::ends_with("_human")
    ) %>%
    dplyr::select(-dplyr::matches("DG_human"))

  out_wide <- file.path(out_dir, sprintf("%s_wide_data_%s.csv", out_prefix, sal_tag))
  utils::write.csv(data_wide_na_rm_reordered, out_wide, row.names = FALSE)

  # 11) average humans
  humans_averaged <- data_wide_na_rm_reordered %>%
    dplyr::mutate(
      mean_human_severity   = rowMeans(dplyr::select(., dplyr::matches("_human")), na.rm = TRUE),
      median_human_severity = apply(dplyr::select(., dplyr::matches("_human")), 1, median, na.rm = TRUE),
      sd_human_severity     = apply(dplyr::select(., dplyr::matches("_human")), 1, sd, na.rm = TRUE),
      n_humans              = rowSums(!is.na(dplyr::select(., dplyr::matches("_human"))))
    ) %>%
    tidyr::drop_na()

  list(
    saliency_col    = sal_col,
    sal_tag         = sal_tag,
    long_df         = df2,
    wide_grid       = data_wide,
    wide_clean      = data_wide_na_rm_reordered,
    humans_averaged = humans_averaged,
    out_grid_csv    = out_grid,
    out_wide_csv    = out_wide
  )
}

```

##### Do not touch the function below, it was painful to make.  ##### 

```{r}
# -----------------------------
# 2) Run “rest of code chunks” on averaged table
# -----------------------------
analyze_human_vs_model <- function(humans_averaged, out_dir = ".", sal_tag = "saliency") {
  df_plot <- humans_averaged %>%
    dplyr::select(-dplyr::any_of(c("n_humans", "sd_human_severity")))
  
  # --- robust numeric selection for pairs() ---
  num_df <- df_plot %>%
    dplyr::select(-dplyr::any_of("filename")) %>%
    dplyr::mutate(dplyr::across(dplyr::everything(), ~ suppressWarnings(as.numeric(.x)))) %>%
    dplyr::select(dplyr::where(is.numeric))

  pdf_path <- file.path(out_dir, sprintf("pairs_plot_human_vs_model_%s.pdf", sal_tag))

  if (ncol(num_df) < 2) {
    warning("Not enough numeric columns for pairs plot; skipping: ", pdf_path)
  } else {
    # --- robust PDF device: try cairo_pdf, else base pdf ---
    opened <- FALSE
    try({
      if ("cairo_pdf" %in% getNamespaceExports("grDevices")) {
        grDevices::cairo_pdf(pdf_path, width = 12, height = 12)  # no family arg here
      } else {
        grDevices::pdf(pdf_path, width = 12, height = 12, family = "sans")
      }
      opened <- TRUE

      pairs(
        num_df,
        upper.panel = NULL,
        lower.panel = lower.panel
      )
    }, silent = TRUE)

    if (opened) grDevices::dev.off()
  }
  
  df_dat <- humans_averaged %>%
  dplyr::select(filename, dplyr::everything())


  human_cols <- names(df_dat) %>% stringr::str_subset("^Human_Rater\\d+$")
  if (length(human_cols) == 0) human_cols <- names(df_dat) %>% stringr::str_subset("_human$")
  
  if (length(human_cols) == 0) {
    stop("No human rater columns found. Columns: ", paste(names(df_dat), collapse = ", "))
  }
  
  # Ensure Mean_Human_Rating exists
  if (!("Mean_Human_Rating" %in% names(df_dat))) {
    df_dat <- df_dat %>%
      dplyr::mutate(
        Mean_Human_Rating = rowMeans(dplyr::across(dplyr::all_of(human_cols)), na.rm = TRUE)
      )
  }
  
 
  # CV candidates by name (not by type)
  cv_candidates <- setdiff(
    names(df_dat),
    c("filename", human_cols, "Mean_Human_Rating", "Median_Human_Rating", "sd_human_severity", "n_humans")
  )
  
  # Coerce CV candidates safely to numeric
  df_dat <- df_dat %>%
    dplyr::mutate(dplyr::across(dplyr::all_of(cv_candidates), ~ suppressWarnings(as.numeric(.x))))
  
  cv_cols <- df_dat %>%
    select(-all_of(human_cols), -filename) %>%
    select(where(is.numeric)) %>%
    names() %>%
    .[. != "median_human_severity" & . != "mean_human_severity" & . != "sd_human_severity" & . != "n_humans" & . != "Mean_Human_Rating"  & . != "Median_Human_Rating"]  

  # CV correlations vs Mean_Human_Rating (robust)
  cv_correlations <- tibble(model = cv_cols) %>%
    mutate(
      correlation = purrr::map_dbl(model, function(col) {
        x <- suppressWarnings(as.numeric(df_dat[[col]]))
        y <- suppressWarnings(as.numeric(df_dat$Mean_Human_Rating))
  
        ok <- is.finite(x) & is.finite(y)
        if (sum(ok) < 3) return(NA_real_)           # not enough pairs
        if (sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NA_real_)  # constant vector
  
        cor(x[ok], y[ok])
      })
    ) %>%
    arrange(desc(abs(correlation)))
  
  cat("n human cols:", length(human_cols), "\n")
  print(human_cols)
 #cat("#######################################################################","\n")

  cat("n cv cols:", length(cv_cols), "\n")
  print(cv_cols)

  
  cat("########################  Performing ICC  ##############################", "\n")
  
  # ICC humans
  icc_out_human <- if (length(human_cols) >= 2) psych::ICC(df_dat %>% select(all_of(human_cols))) else NA
  icc_out_cv    <- if (length(cv_cols)   >= 2) psych::ICC(df_dat %>% select(all_of(cv_cols)))   else NA
  
  cat("[", sal_tag, "] ICC results:\n")
  print(icc_out_human)

  # variance components: humans
  df <- humans_averaged %>% dplyr::select(filename, dplyr::everything())
  
  # Human columns: prefer Human_Rater*, else fall back to *_human
  human_cols <- names(df) %>% stringr::str_subset("^Human_Rater\\d+$")
  if (length(human_cols) == 0) {
    human_cols <- names(df) %>% stringr::str_subset("_human$")
  }
  
  # Fail fast with a clear message (instead of pivot_longer cryptic error)
  if (length(human_cols) == 0) {
    stop(
      "No human rater columns found in humans_averaged.\n",
      "Expected columns like Human_Rater1.. or *_human.\n",
      "Columns present: ", paste(names(df), collapse = ", ")
    )
  }
  
  # If Mean_Human_Rating doesn't exist (renaming didn’t happen), compute it now
  if (!("Mean_Human_Rating" %in% names(df))) {
    df <- df %>%
      dplyr::mutate(Mean_Human_Rating = rowMeans(dplyr::across(dplyr::all_of(human_cols)), na.rm = TRUE))
  }
  
  # Now pivot safely
  df_long_humans <- df %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(human_cols),
      names_to = "rater",
      values_to = "human_score"
    )
  
  cat("\n")
  cat("#######################  Fitting linear model  #########################", "\n")


  m_human_vc <- lmer(human_score ~ 1 + (1|filename) + (1|rater), data = df_long_humans, REML = TRUE)
  vc <- as.data.frame(VarCorr(m_human_vc))
  var_image <- vc %>% filter(grp == "filename") %>% pull(vcov)
  var_rater <- vc %>% filter(grp == "rater")    %>% pull(vcov)
  var_resid <- attr(VarCorr(m_human_vc), "sc")^2
  total_var <- var_image + var_rater + var_resid

  vc_tbl <- tibble(
    component = c("image", "rater", "residual"),
    variance  = c(var_image, var_rater, var_resid),
    pct       = 100 * variance / total_var
  )
  
  cat("[", sal_tag, "] attributable variance:\n")
  print(vc_tbl)

  #print(df)
  #class(df)
  #is.function(df)
  
  # CV vs mean human metrics
  df_human_summary <- df %>%
    mutate(
      human_mean = rowMeans(across(all_of(human_cols)), na.rm = TRUE),
      human_sd   = apply(select(., all_of(human_cols)), 1, sd, na.rm = TRUE)
    )

  cv_vs_mean_tbl <- map_dfr(cv_cols, \(cv) {
    out <- metric_vs_reference(df_human_summary[[cv]], df_human_summary$human_mean)
    out$model <- cv
    out
  }) %>% relocate(model)

    cat("################  Human vs saliency method concordance  ###############", "\n")

  # CV vs each human
  cv_vs_each_human <- expand_grid(model = cv_cols, rater = human_cols) %>%
    mutate(metrics = map2(model, rater, \(m, r) {
      metric_vs_reference(df[[m]], df[[r]])
    })) %>%
    unnest(metrics) #%>%
    #mutate(
    #  CCC  = map_dbl(CCC,  ~ .x[1]),
    #  cor  = map_dbl(cor,  ~ .x[1]),
    #  MAE  = map_dbl(MAE,  ~ .x[1]),
    #  RMSE = map_dbl(RMSE, ~ .x[1]),
    #  bias = map_dbl(bias, ~ .x[1])
    #)

  cv_vs_each_human_summary <- cv_vs_each_human %>%
    dplyr::filter(!rater %in% c("Mean_Human_Rating", "Median_Human_Rating")) %>%
    group_by(model) %>%
    summarize(
      CCC_mean  = mean(CCC,  na.rm = TRUE),
      CCC_sd    = sd(CCC,    na.rm = TRUE),
      MAE_mean  = mean(MAE,  na.rm = TRUE),
      RMSE_mean = mean(RMSE, na.rm = TRUE),
      .groups = "drop"
    )
  
  
  cat("[", sal_tag, "] Concordance summaries:\n")

  print(cv_vs_each_human_summary)
  
  

  # Plots: scatter and BA
  scatter_plot <- wrap_plots(lapply(cv_cols, \(nm) plot_cv_scatter(df_human_summary, nm)))
  ba_plot      <- wrap_plots(lapply(cv_cols, \(nm) plot_ba(df_human_summary, nm)))
  err_plot     <- wrap_plots(lapply(cv_cols, \(nm) cv_error_vs_disagreement(df_human_summary, nm)))

  # Mixed models: human-on-CV (logit)
  models_human_on_cv <- setNames(lapply(cv_cols, \(cv) {
    tryCatch(fit_human_on_cv(df, human_cols, cv), error = function(e) NULL)
  }), cv_cols)
  
  model_summaries <- lapply(models_human_on_cv, function(m) if (is.null(m)) NULL else summary(m))

  
  cat("#########################  Writing files  #############################", "\n")
  cat("[", sal_tag, "] output files saved to:", out_dir, "\n")

  
  # Save key tables to disk (optional)
  write.csv(cv_correlations, file.path(out_dir, sprintf("cv_correlations_%s.csv", sal_tag)), row.names = FALSE)
  write.csv(cv_vs_mean_tbl,  file.path(out_dir, sprintf("cv_vs_mean_tbl_%s.csv", sal_tag)), row.names = FALSE)
  write.csv(cv_vs_each_human_summary, file.path(out_dir, sprintf("cv_vs_each_human_summary_%s.csv", sal_tag)), row.names = FALSE)
  write.csv(vc_tbl, file.path(out_dir, sprintf("human_variance_components_%s.csv", sal_tag)), row.names = FALSE)

  list(
    pdf_pairs = pdf_path,
    human_cols = human_cols,
    cv_cols = cv_cols,
    cv_correlations = cv_correlations,
    icc_humans = icc_out_human,
    icc_cv = icc_out_cv,
    human_vc_model = m_human_vc,
    human_vc_tbl = vc_tbl,
    cv_vs_mean_tbl = cv_vs_mean_tbl,
    cv_vs_each_human = cv_vs_each_human,
    cv_vs_each_human_summary = cv_vs_each_human_summary,
    scatter_plot = scatter_plot,
    ba_plot = ba_plot,
    error_vs_disagreement_plot = err_plot,
    human_on_cv_models = models_human_on_cv,
    human_on_cv_summaries = model_summaries
  )
}
```

## Run the functions above by editing chunk below

```{r}
# 3) Run everything for all saliency methods
# -----------------------------
saliency_cols <- c("Gradient-SG_sr2", "GradCAM_sr2", "DeepLift_sr2")

# optional: where to save outputs
out_dir <- "../figs_for_pub"

results <- setNames(lapply(saliency_cols, function(col) {
  prep <- make_human_vs_model(
    correlation_matrix_human_v_model,
    concatenated_df_trimmed,
    saliency_col = col,
    out_dir = out_dir
  )

  analysis <- analyze_human_vs_model(
    prep$humans_averaged,
    out_dir = out_dir,
    sal_tag = col
  )

  list(prep = prep, analysis = analysis)
}), saliency_cols)

# -----------------------------
# 4) Example: look at correlations for each method
# -----------------------------
results[["GradCAM_sr2"]]$analysis$cv_correlations %>% slice(1:20)
results[["Gradient-SG_sr2"]]$analysis$cv_correlations %>% slice(1:20)


# Example: view plots for one method in RStudio
results[["GradCAM_sr2"]]$analysis$scatter_plot
results[["GradCAM_sr2"]]$analysis$error_vs_disagreement_plot
results[["Gradient-SG_sr2"]]$analysis$scatter_plot
results[["Gradient-SG_sr2"]]$analysis$error_vs_disagreement_plot
results[["DeepLift_sr2"]]$analysis$scatter_plot
results[["DeepLift_sr2"]]$analysis$error_vs_disagreement_plot


# Example: inspect ICC outputs
results[["GradCAM_sr2"]]$analysis$icc_humans
results[["Gradient-SG_sr2"]]$analysis$icc_humans
results[["DeepLift_sr2"]]$analysis$icc_humans


# Example: summaries of mixed models (human-on-CV) for one method
results[["GradCAM_sr2"]]$analysis$human_on_cv_summaries
results[["Gradient-SG_sr2"]]$analysis$human_on_cv_summaries
results[["DeepLift_sr2"]]$analysis$human_on_cv_summaries


results[["GradCAM_sr2"]]$analysis$cv_vs_each_human_summary
results[["Gradient-SG_sr2"]]$analysis$cv_vs_each_human_summary
results[["DeepLift_sr2"]]$analysis$cv_vs_each_human_summary



```
# ------------------------------


# What about human estimates of sporulation? 

```{r}

# select only the relevant columns
sporulation_select <- combined %>%
        dplyr::select(filename, identifier, model_type, tray, classes, up_th, down_th, spor_th, inf_gate, sal_threshold, percent_conidiophores, percent_hyphae) %>%
        dplyr::filter(!is.na(percent_conidiophores)) %>%
        dplyr::filter(tray == "10dpi")


# if the filename doesnt end in "5dpi" or "10dpi", then append this information to the filename from the tray column
sporulation_select <- sporulation_select %>%
  dplyr::mutate(filename = dplyr::case_when(
    grepl("10dpi", filename) ~ filename,
    TRUE ~ paste0(filename, "_", tray)
  ))

sporulation_select2 <- sporulation_select %>%
  dplyr::mutate(
    up_th_num = suppressWarnings(as.numeric(up_th)),
    identifier = dplyr::if_else(
      !is.na(up_th_num),                      # only if it's numeric
      paste0(identifier, "_up", up_th),         # append original value
      identifier                              # otherwise leave unchanged
    )
  ) %>%
  dplyr::mutate(
    down_th_num = suppressWarnings(as.numeric(down_th)),
    identifier = dplyr::if_else(
      !is.na(down_th_num),                      # only if it's numeric
      paste0(identifier, "_dwn", down_th),         # append original value
      identifier                              # otherwise leave unchanged
      )) %>%
    dplyr::mutate(
    inf_gate_num = suppressWarnings(as.numeric(inf_gate)),
    identifier = dplyr::if_else(
      !is.na(inf_gate_num),                      # only if it's numeric
      paste0(identifier, "_infg", inf_gate),         # append original value
      identifier                              # otherwise leave unchanged
      )) %>%
    dplyr::mutate(
      spor_th_num = suppressWarnings(as.numeric(spor_th)),
      identifier = dplyr::if_else(
      !is.na(spor_th_num),                      # only if it's numeric
      paste0(identifier, "_spth", spor_th),         # append original value
      identifier                              # otherwise leave unchanged
      )) %>%
    dplyr::mutate(
      sal_th_num = suppressWarnings(as.numeric(sal_threshold)),
      identifier = dplyr::if_else(
        !is.na(sal_th_num),                      # only if it's numeric
        paste0(identifier, "_salth", sal_threshold),         # append original value
        identifier                             # otherwise leave unchanged
        )) %>%
dplyr::select(c(-up_th_num, -down_th_num, -inf_gate_num, -spor_th_num, -sal_th_num))

unique(sporulation_select2$identifier)


sporulation_select2 <- sporulation_select2 %>%
  dplyr::select(filename, identifier, model_type, classes, percent_conidiophores, percent_hyphae)

write.csv(sporulation_select2, "human_vs_model_long_sporulation_data.csv")


sporulation_select3 <- sporulation_select2 %>%
  dplyr::filter(identifier %in% c("MW",
                                 "DG",
                                 "NW",
                                 "TB",
                                 "BR",
                                 "CC",
                                 "DG",
                                 "Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5",
                                 "Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.7_salth0.7",
                                 "Jan26_23-15-35_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5",
                                 "Jan26_23-15-35_2026_up0.95_dwn0.2_infg0.3_spth0.5_salth0.7",
                                 "Feb03_16-24-20_2024_up0.8_dwn0.3_salth0.7",
                                 "Feb14_15-53-04_2024_up0.8_dwn0.3_salth0.7",
                                 "Jan29_22-02-46_2026_up0.55_dwn0.15_infg0.3_spth0.65_salth0.5"
                                 ))

unique(sporulation_select2$identifier)

library(data.table)

tmp <- sporulation_select3 %>%
  mutate(
    percent_conidiophores = readr::parse_number(
      na_if(as.character(percent_conidiophores), "-")
    )
  ) %>%
    mutate(
    percent_hyphae = readr::parse_number(
      na_if(as.character(percent_hyphae), "-")
    )
  )

#dt <- as.data.table(tmp)

#dt_nodups <- dt[, if (.N == 1) .SD, by = .(filename, identifier, model_type)]


# pivot the table
data_spore_wide <- dcast(
  as.data.table(tmp),
  filename ~ identifier + model_type,
  value.var = "percent_conidiophores",
  fun.aggregate = mean,
  na.rm = TRUE
)

data_hyphae_wide <- dcast(
  as.data.table(tmp),
  filename ~ identifier + model_type,
  value.var = "percent_hyphae",
  fun.aggregate = mean,
  na.rm = TRUE
)

##data_wide <- dcast(
#  dt_nodups,
#  filename ~ identifier + model_type,
#  value.var = "severity_rate_patch"
#)

write.csv2(data_spore_wide, "human_vs_model_wide_data_sporulation_10dpi.csv")

# drop columns that start with Feb01_ or Aug31_ *these are old model versions that we dont want to include in the analysis*
data_wide_spore_na_rm <- data_spore_wide %>%
  dplyr::select(-dplyr::matches("^(Feb01_|Aug31_|Feb03)")) %>%
  dplyr::select(-dplyr::matches("May11_00-29-56_2021")) %>% # remove this specific column 
  dplyr::filter(!is.na(TB_human))


data_wide_na_rm_spore_reordered <- data_wide_spore_na_rm %>%
  dplyr::select(
    -dplyr::ends_with("_computer_vision"),
    -dplyr::ends_with("_human"),
    dplyr::ends_with("_computer_vision"),
    dplyr::ends_with("_human")
  )


write.csv(data_wide_na_rm_spore_reordered, "human_vs_model_wide_spore.csv")


# drop Dave since he didnt recieve great instructions
data_wide_na_rm_spore_reordered <- data_wide_na_rm_spore_reordered %>%
  dplyr::select(-dplyr::matches("DG_human"))
  

# Average out the human ratings
data_wide_na_rm_reordered_humans_spore_averaged <- data_wide_na_rm_spore_reordered %>%
  dplyr::mutate(
    mean_human_severity   = rowMeans(dplyr::select(., dplyr::matches("_human")), na.rm = TRUE),
    median_human_severity = apply(dplyr::select(., dplyr::matches("_human")), 1, median, na.rm = TRUE),
    sd_human_severity     = apply(dplyr::select(., dplyr::matches("_human")), 1, sd, na.rm = TRUE),
    n_humans              = rowSums(!is.na(dplyr::select(., dplyr::matches("_human"))))
  ) %>%
  # remove any NaN
  drop_na()

# rename BR_Human to Human_Rater1; CC_Human to Human_Rater2; MW_Human to Human_Rater3; NW_Human to Human_Rater4; TB_Human to Human_Rater5
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("BR_human", "Human_Rater1", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("CC_human", "Human_Rater2", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("MW_human", "Human_Rater3", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("NW_human", "Human_Rater4", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("TB_human", "Human_Rater5", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("mean_human_severity", "Mean_Human_Rating", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("median_human_severity", "Median_Human_Rating", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.7_salth0.7_computer_vision", "Jan22_10-21-00_2026_sal0.7", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("Jan26_23-15-35_2026_up0.95_dwn0.2_infg0.3_spth0.5_salth0.7_computer_vision", "Jan26_23-15-35_2026_sal0.7", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("Jan26_23-15-35_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5_computer_vision", "Jan26_23-15-35_2026_sal0.5", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("Jan22_10-21-00_2026_up0.95_dwn0.3_infg0.3_spth0.5_salth0.5_computer_vision", "Jan22_10-21-00_2026_sal0.5", colnames(data_wide_na_rm_reordered_humans_spore_averaged))
colnames(data_wide_na_rm_reordered_humans_spore_averaged) <- gsub("Jan29_22-02-46_2026_up0.55_dwn0.15_infg0.3_spth0.65_salth0.5", "Jan29_22-02-46_2026_sal0.5", colnames(data_wide_na_rm_reordered_humans_spore_averaged))


```

# Sporulation condordance and correlation

```{r}


lower.panel <- function(x, y){
  points(x, y, pch = 19)

  r <- round(cor(x, y), 2)
  txt <- paste0("R = ", r)

  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))

  x0 <- 0.5
  y0 <- 0.9
  off <- 0.01     # outline thickness
  cex_txt <- 1.5   # text size
  outline_col <- "black"

  # 8-direction outline offsets
  offsets <- matrix(c(
    -off,  0,
     off,  0,
     0,   -off,
     0,    off,
    -off, -off,
    -off,  off,
     off, -off,
     off,  off
  ), ncol = 2, byrow = TRUE)

  # draw outline
  for(i in 1:nrow(offsets)){
    text(x0 + offsets[i, 1], y0 + offsets[i, 2],
         txt, col = outline_col, font = 2, cex = cex_txt)
  }

  # draw main text on top
  text(x0, y0, txt, col = "#00E5FF", font = 2, cex = cex_txt)
}


df_plot <- data_wide_na_rm_reordered_humans_spore_averaged %>%
  select(-n_humans, -sd_human_severity)   # keep filename, drop stats you don't want

library(Cairo)

CairoPDF("../figs_for_pub/correlation_plots/pairs_plot_human_vs_model_grid_sporulation.pdf",
         width = 12, height = 12,
         family = "Arial")

pairs(
  df_plot %>% select(-filename) %>% select(where(is.numeric)),
  upper.panel = NULL,
  lower.panel = lower.panel
)

dev.off()

# 1) Get only the computer vision columns
cv_cols <- names(data_wide_na_rm_reordered_humans_spore_averaged) %>%
  .[grepl("_202", .)]

cv_cols

# 2) Compute correlation vs mean_human_severity for each CV model
cv_correlations <- tibble(model = cv_cols) %>%
  mutate(
    correlation = sapply(model, function(col) {
      cor(
        data_wide_na_rm_reordered_humans_spore_averaged[[col]],
        data_wide_na_rm_reordered_humans_spore_averaged$Mean_Human_Rating,
        use = "pairwise.complete.obs"
      )
    })
  ) %>%
  arrange(desc(abs(correlation)))  # sort by strongest correlation (positive or negative)

# 3) Print top 5
cv_correlations %>% slice(1:50)



#### Human inter-rater reliability (ICC) ######
df_spore <- data_wide_na_rm_reordered_humans_spore_averaged %>%
  select(filename, starts_with("Human_Rater"), everything())


human_cols_spore <- names(df_spore) %>% str_subset("^Human_Rater\\d+$")

# If your 3 CV models are exactly those date columns, grab all non-human numeric columns except filename:
cv_cols_spore <- df_spore %>%
  select(-all_of(human_cols_spore), -filename) %>%
  select(where(is.numeric)) %>%
  names() %>%
  .[. != "Mean_Human_Rating" & . != "Median_Human_Rating" & . != "sd_human_severity" & . != "n_humans"]

human_cols_spore
cv_cols_spore


# Calculate ICC for human raters
human_mat_spore <- df_spore %>% select(all_of(human_cols_spore))

# ICC(2,1) and ICC(2,k) are common when raters are considered a random sample
icc_out_spore <- psych::ICC(human_mat_spore)

print(icc_out_spore)

# Calculate ICC for CV models (not that informative tbh)
cv_mat_spore <- df_spore %>% select(all_of(cv_cols_spore))
icc_out_cv_spore <- psych::ICC(cv_mat_spore)
print(icc_out_cv_spore)


df_long_humans_spore <- df_spore %>%
  pivot_longer(cols = all_of(human_cols_spore), names_to = "rater", values_to = "human_score")


m_human_vc_spore <- lmer(human_score ~ 1 + (1|filename) + (1|rater), data = df_long_humans_spore, REML = TRUE)
summary(m_human_vc_spore)

# Variance components for the spore model
vc_human_spore <- as.data.frame(VarCorr(m_human_vc_spore))

var_image_spore <- vc_human_spore %>% filter(grp == "filename") %>% pull(vcov)
var_rater_spore <- vc_human_spore %>% filter(grp == "rater")    %>% pull(vcov)

# Residual variance: use the SAME model object
var_resid_spore <- attr(VarCorr(m_human_vc_spore), "sc")^2

total_var_spore <- var_image_spore + var_rater_spore + var_resid_spore

tibble(
  component = c("image", "rater", "residual"),
  variance  = c(var_image_spore, var_rater_spore, var_resid_spore),
  pct       = 100 * variance / total_var_spore
)



df_human_summary_spore <- df_spore %>%
  mutate(
    human_mean = rowMeans(across(all_of(human_cols_spore)), na.rm = TRUE),
    human_sd   = apply(select(., all_of(human_cols_spore)), 1, sd, na.rm = TRUE)
  )

cv_vs_mean_tbl_spore <- purrr::map_dfr(cv_cols_spore, \(cv) {
  out <- metric_vs_reference(
    dplyr::pull(df_human_summary_spore, all_of(cv)),
    df_human_summary_spore$human_mean
  )
  out$model <- cv
  out
}) %>%
  dplyr::relocate(model)


cv_vs_mean_tbl_spore


cv_vs_each_human_spore <- expand_grid(model = cv_cols_spore, rater = human_cols_spore) %>%
  mutate(metrics = map2(model, rater, \(m, r) {
    metric_vs_reference(df_spore[[m]], df_spore[[r]])
  })) %>%
  unnest(metrics)


cv_vs_each_human_spore2 <- cv_vs_each_human_spore %>%
  mutate(
    CCC  = map_dbl(CCC,  ~ .x[1]),
    cor  = map_dbl(cor,  ~ .x[1]),
    MAE  = map_dbl(MAE,  ~ .x[1]),
    RMSE = map_dbl(RMSE, ~ .x[1]),
    bias = map_dbl(bias, ~ .x[1])
  )

cv_vs_each_human_spore2 %>%
  dplyr::filter(!rater %in% c("Mean_Human_Rating", "Median_Human_Rating")) %>%
  group_by(model) %>%
  summarize(
    CCC_mean = mean(CCC, na.rm = TRUE),
    CCC_sd   = sd(CCC,  na.rm = TRUE),
    MAE_mean = mean(MAE, na.rm = TRUE),
    RMSE_mean= mean(RMSE, na.rm = TRUE),
    .groups = "drop"
  )




```

Wow, raters really closely agree on sporulation percentage (92% variance explained by image, only 1% by rater). 

Pearson r corr for Jan26 model: 0.9815904 and CCC: 0.9803512

# ------------------------------


## Histogram distributions of human ratings


```{r, histogram of human ratings, warning=FALSE}

dist_long <- data_wide_na_rm_reordered_humans_averaged %>%
  mutate(
    dpi = str_extract(filename, "(?<=_)\\d+(?=dpi$)"),
    dpi = factor(paste0(dpi, " dpi"), levels = c("5 dpi", "10 dpi"))
  ) %>%
  select(
    dpi,
    Mean_Human_Rating,
    `Jan26_23-15-35_2026_sal0.5`
  ) %>%
  pivot_longer(
    cols = c(Mean_Human_Rating, `Jan26_23-15-35_2026_sal0.5`),
    names_to = "source",
    values_to = "severity"
  ) %>%
  filter(!is.na(dpi), !is.na(severity)) %>%
  mutate(
    source = recode(
      source,
      Mean_Human_Rating = "Human (mean)",
      `Jan26_23-15-35_2026_sal0.5` = "Computer vision"
    )
  )

ggplot(dist_long, aes(x = severity, fill = source)) +
  geom_density(alpha = 0.5, adjust = 1.2) +
  facet_wrap(~ dpi, nrow = 1, scales = "fixed") +
  labs(
    title = "Distribution of Disease Severity by Time Point",
    x = "Disease severity (%)",
    y = "Density",
    fill = "Rating source"
  ) +
  theme_classic()



```
## Segmentation correlation

Trained a very quick and dirty one class DeepLab segmentation model (~200 training, ~40 testing) to see how it would compare. This isnt a fair comparison because obviously there is a lot more classification data, but segmentation labeling takes ages longer (up to 20 minutes per patch) so I dont think a segmentation approach is feasible at this point (especially given the classification model already does so well).

```{r look_at_segmentation_correlation, message=FALSE, warning=FALSE}

# 1) read in csv to segmentation_results
#seg_path <- "C:/Users/Intel User/Desktop/blackbird_scripts/results/human_disk_assessments/segmentation/DeepLab-Feb04_00-32-07_2026/5dpi_and_10dpi_inference_summary.csv"

seg_path <- "../spreadsheets/segmentation_results.csv"
segmentation_results <- read_csv(seg_path, show_col_types = FALSE)
#write.csv(segmentation_results, "../spreadsheets/segmentation_results.csv")

# 2) pull just what we need, and de-duplicate filenames if necessary
seg_lookup <- segmentation_results %>%
  dplyr::select(filename, pct_infected_of_leaf) %>%
  group_by(filename) %>%
  summarise(pct_infected_of_leaf = mean(pct_infected_of_leaf, na.rm = TRUE), .groups = "drop")

# 3) match filenames and add as a new column in your wide human dataset
new_col_name <- "DeepLab-Feb04_00-32-07_2026"

# Uncomment these if you want to include Dave
# making this explicit for easy intelligibility
#data_wide_na_rm_reordered_with_dg <- data_wide_na_rm %>%
#  dplyr::select(
#    -dplyr::ends_with("_computer_vision"),
#    -dplyr::ends_with("_human"),
#    dplyr::ends_with("_computer_vision"),
#    dplyr::ends_with("_human")
#  )

# Average out the human ratings
#data_wide_na_rm_reordered_humans_averaged_with_dg <- data_wide_na_rm_reordered_with_dg %>%
#  dplyr::mutate(
#    mean_human_severity   = rowMeans(dplyr::select(., dplyr::matches("_human")), na.rm = TRUE),
#    median_human_severity = apply(dplyr::select(., dplyr::matches("_human")), 1, median, na.rm = TRUE),
#    sd_human_severity     = apply(dplyr::select(., dplyr::matches("_human")), 1, sd, na.rm = TRUE),
#    n_humans              = rowSums(!is.na(dplyr::select(., dplyr::matches("_human"))))
#  ) %>%
#  # remove any NaN
#  drop_na()

#data_wide_na_rm_reordered_humans_averaged_with_seg <- data_wide_na_rm_reordered_humans_averaged_with_dg %>%
#  left_join(seg_lookup, by = "filename") %>%
#  rename(!!new_col_name := pct_infected_of_leaf)

# this is without Dave since he didnt recieve good instructions
data_wide_na_rm_reordered_humans_averaged_with_seg <- data_wide_na_rm_reordered_humans_averaged %>%
   left_join(seg_lookup, by = "filename") %>%
   rename(!!new_col_name := pct_infected_of_leaf)

# also with the sporulation dataset
data_wide_na_rm_reordered_humans_spore_averaged_with_seg <- data_wide_na_rm_reordered_humans_spore_averaged %>%
   left_join(seg_lookup, by = "filename") %>%
   rename(!!new_col_name := pct_infected_of_leaf)

# 4) make correlation matrices (numeric columns only)
numeric_df_severity <- data_wide_na_rm_reordered_humans_averaged_with_seg %>%
  na.omit()
  
df_plot_segmentation <- numeric_df_severity %>%
  dplyr::select(-n_humans, -sd_human_severity) %>%
  dplyr::select(where(is.numeric)) %>%
  na.omit()
  
colnames(df_plot_segmentation) <- substr(colnames(df_plot_segmentation), 1, 12)
  
CairoPDF("../figs_for_pub/correlation_plots/pairs_plot_human_vs_model_segmentation.pdf",
           width = 12, height = 12,
           family = "Arial")
  
labels_12 <- substr(colnames(df_plot_segmentation), 1, 12)

pairs(
  df_plot_segmentation,
  labels = labels_12,
  upper.panel = NULL,
  lower.panel = lower.panel,
  cex.labels = 1.4
)
  
  dev.off()
  
  
  ##############################
  # Print out top correlations #
  ##############################

  
# 1) Get only the computer vision columns
cv_cols_seg <- names(df_plot_segmentation) %>%
  .[grepl("-", .)]


# 2) Compute correlation vs mean_human_severity for each CV model
cv_correlations_seg <- tibble(model = cv_cols_seg) %>%
  mutate(
    correlation = sapply(model, function(col) {
      cor(
        df_plot_segmentation[[col]],
        df_plot_segmentation$Mean_Human_R,
        use = "pairwise.complete.obs"
      )
    })
  ) %>%
  arrange(desc(abs(correlation)))  # sort by strongest correlation (positive or negative)

# 3) Print top 5
cv_correlations_seg %>% slice(1:50)

# 200 leaf disk set with n = 5 raters
# Jan26_23-15-	0.9692613			
# Jan29_22-02-	0.9676133			
# DeepLab-Feb0	0.8190004


##############################
#        Concordance         #
##############################

cv_cols_seg <- names(numeric_df_severity) %>%
  .[grepl("-", .)] 

human_cols_seg <- grep("Human_", names(numeric_df_severity), value = TRUE) 

human_cols_seg_no_mean <- setdiff(
  human_cols_seg,
  c("Mean_Human_Rating", "Median_Human_Rating")
)

# without summary metrics, clean
cv_vs_mean_tbl_seg <- purrr::map_dfr(human_cols_seg_no_mean, \(cv) {
  out <- metric_vs_reference(
    dplyr::pull(numeric_df_severity, all_of(cv)),
    numeric_df_severity$Mean_Human_Rating
  )
  out$model <- cv
  out
}) %>%
  dplyr::relocate(model)

cat("Mean of human CCC with each other (n=5):\n") 
mean(cv_vs_mean_tbl_seg$CCC)

# kept the mean and median in this one
cv_vs_each_human_seg_with_summary_metrics <- expand_grid(model = cv_cols_seg, rater = human_cols_seg) %>%
  dplyr::mutate(metrics = map2(model, rater, \(m, r) {
    metric_vs_reference(numeric_df_severity[[m]], numeric_df_severity[[r]])
  })) %>%
  unnest(metrics)



cv_vs_each_human_seg_with_summary_metrics2 <- cv_vs_each_human_seg_with_summary_metrics %>%
  mutate(
    CCC  = map_dbl(CCC,  ~ .x[1]),
    cor  = map_dbl(cor,  ~ .x[1]),
    MAE  = map_dbl(MAE,  ~ .x[1]),
    RMSE = map_dbl(RMSE, ~ .x[1]),
    bias = map_dbl(bias, ~ .x[1])
  )

cv_vs_each_human_seg_with_summary_metrics2 %>%
  dplyr::filter(!rater %in% c("Mean_Human_Rating", "Median_Human_Rating")) %>%
  group_by(model) %>%
  summarize(
    CCC_mean = mean(CCC, na.rm = TRUE),
    CCC_sd   = sd(CCC,  na.rm = TRUE),
    MAE_mean = mean(MAE, na.rm = TRUE),
    RMSE_mean= mean(RMSE, na.rm = TRUE),
    .groups = "drop"
  )

```
Okay so mean consensus concordance with J26 model is 0.9245640 in humans (vs mean human rating CCC with J26 model is 0.9678170). Note the difference -- for the former we take the mean after vs the later we take the mean before. 

For DeepLab Feb4: 0.2454396 vs 0.2543258


# ---------------------------------
# Redo saliency correlations with optimized IoU saliency thresholds


```{r saliency optimized thresholds}

optimized_saliency_results_path <- "../spreadsheets/human_vs_bb_models_disase_severity.csv"
final_joined <- read_csv(optimized_saliency_results_path, show_col_types = FALSE)

##### okay, loaded the and appended the J26_fixed_0.5 (across both head) sal data ##### 
# technically they're all fixed thresholded but the 0.5 ones are labeled 05_fixed and they're fixed to 0.5 across both heads

cv_cols <- names(final_joined)[
  grepl("J26", names(final_joined)) | grepl("DeepLab", names(final_joined))
]


human_cols <- grep("Human_", names(final_joined), value = TRUE) 

human_cols_no_mean <- setdiff(
  human_cols,
  c("Mean_Human_Rating", "Median_Human_Rating")
)


# kept the mean and median in this one
cv_and_sal_vs_each_human_with_summary_metrics <- expand_grid(model = cv_cols, rater = human_cols) %>%
  dplyr::mutate(metrics = map2(model, rater, \(m, r) {
    metric_vs_reference(final_joined[[m]], final_joined[[r]])
  })) %>%
  unnest(metrics)


cv_and_sal_vs_each_human_with_summary_metrics

cv_and_sal_vs_each_human_with_summary_metrics2 <- cv_and_sal_vs_each_human_with_summary_metrics %>%
  mutate(
    CCC  = map_dbl(CCC,  ~ .x[1]),
    cor  = map_dbl(cor,  ~ .x[1]),
    MAE  = map_dbl(MAE,  ~ .x[1]),
    RMSE = map_dbl(RMSE, ~ .x[1]),
    bias = map_dbl(bias, ~ .x[1])
  )

cv_and_sal_vs_each_human_with_summary_metrics2 %>%
  dplyr::filter(!rater %in% c("Mean_Human_Rating", "Median_Human_Rating")) %>%
  group_by(model) %>%
  summarize(
    CCC_mean = mean(CCC, na.rm = TRUE),
    CCC_sd   = sd(CCC,  na.rm = TRUE),
    MAE_mean = mean(MAE, na.rm = TRUE),
    RMSE_mean= mean(RMSE, na.rm = TRUE),
  )

cv_and_sal_vs_each_human_with_summary_metrics2 %>%
  dplyr::filter(rater == "Mean_Human_Rating")

write.csv(
  cv_and_sal_vs_each_human_with_summary_metrics2,
  here::here("..", "spreadsheets", "human_vs_bb_models_summary_metrics.csv"),
  row.names = FALSE
)

######### Correlation


final_joined_select <- final_joined %>%
  dplyr::select(-n_humans, -sd_human_severity) %>%
  dplyr::select(where(is.numeric)) %>%
  na.omit()
  
colnames(final_joined_select) <- substr(colnames(final_joined_select), 1, 20)
  
CairoPDF("../figs_for_pub/correlation_plots/pairs_plot_human_vs_model_optimal_saliency_thresholds.pdf",
           width = 18, height = 18,
           family = "Arial")
  
labels_12 <- substr(colnames(final_joined_select), 1, 20)

pairs(
  final_joined_select,
  labels = labels_12,
  upper.panel = NULL,
  lower.panel = lower.panel,
  cex.labels = 0.8
)
  
  dev.off()

  
  
# concordance between models? 

cv_and_sal_vs_cv_sal <- expand_grid(model = cv_cols, rater = cv_cols) %>%
dplyr::mutate(metrics = map2(model, rater, \(m, r) {
  metric_vs_reference(final_joined[[m]], final_joined[[r]])
})) %>%
unnest(metrics)


write.csv(
  cv_and_sal_vs_cv_sal,
  here::here("..", "spreadsheets", "model_vs_models_summary_metrics.csv"),
  row.names = FALSE
)

```

correlation with deeplab segmentation model using IoU optimized sal-threshold: 
gradcam: 0.96
deeplift: 0.95
smoothgrad: 0.95

correlation with deeplab segmentation model when using fixed 0.5 sal threshold (both heads): 
gradcam 0.81
deeplift: 0.84
smoothgrad 0.86


concordance with deeplab segmentation model using IoU optimized sal-threshold: 
GradCAM 0.82
DeepLift 0.90
SmoothGrad 0.63

concordance with deeplab segmentation when using fixed 0.5 sal threshold (both heads): 
GradCAM 0.76
DeepLIFT 0.24
SmoothGrad 0.49


\begin{table}[ht]
\centering
\caption[Post-inference saliency map performance of the best-performing model with and without optimized fixed intersection-over-union (IoU) thresholds.]
{\textbf{Post-inference saliency map performance of the best-performing model with and without optimized fixed intersection-over-union (IoU) thresholds.}. 
Pearson’s $r$ reflects the strength of linear association with the mean human ratings, while the CCC captures absolute agreement by accounting for both correlation and systematic bias. MAE = mean absolute error; RMSE = root mean square error. }
\label{tab:saliency_jan26}
\begin{tabular}{lccccccc}
\toprule
Method & Head 1 Threshold & Head 2 Threshold & Pearson $r$ & CCC & MAE & RMSE & Bias \\
\midrule
DeepLIFT              & 0.5 & 0.5 & 0.96 & 0.05 & 25.11 & 37.90 & -25.10        \\
Gradient + SmoothGrad & 0.5 & 0.5 & 0.97 & 0.11 & 24.05 & 36.18 & -24.03       \\
GradCAM               & 0.5 & 0.5 & 0.96 & 0.40 & 18.68 & 28.16 & -18.59        \\
DeepLIFT              & 0.4 & 0.1 & 0.84 & 0.34 & 20.73 & 30.44 & -20.70 \\
Gradient + SmoothGrad & 0.5 & 0.1 & 0.82 & 0.58 & 16.90 & 24.98 & -16.85 \\
GradCAM               & 0.7 & 0.2 & 0.88 & 0.46 & 18.68 & 27.29 & -18.63 \\
\bottomrule
\end{tabular}
\end{table}

